---
title: "Count, PCA and Fingerprint on features"
author: "Florent Chuffart"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
---


```{r header, child="header.Rmd"}
```


```{r}
if (!exists("prefix")) {prefix = "tiles"}
if (!exists("COUNT")) {COUNT = FALSE}
```

- prefix="`r prefix`" is used to reference the files generated by the pipeline.
- `COUNT` = `r COUNT` controls whether the read-counting step using `featureCounts` is performed and is set to `FALSE` by default.


```
prefix="`r prefix`" ; rmarkdown::render("03_count.Rmd", output_file=paste0("03_count_", prefix, ".html"))
```



```{r samples, child="samples.Rmd"}
```

# Features

This section defines the genomic coordinates used throughout the analysis.



```{r features, child=paste0("99_", sub("[0-9]+$", "", prefix), ".Rmd")}
```


## `featureCounts`


```{r featureCounts, results="verbatim"}
# Counts
feat_saf = feat[, c(4,1:3, 6)]
# feat_saf = feat[1:100,c(4,1:3, 6)]
colnames(feat_saf) = c("GeneID", "Chr", "Start", "End", "Strand")
feat_saf$Start = as.numeric(feat_saf$Start)
feat_saf$End   = as.numeric(feat_saf$End  )
saf_filename = paste0(prefix, ".saf")
write.table(feat_saf, file=saf_filename, sep="\t", quote=FALSE,row.names=FALSE, col.names=FALSE)

cmd = "featureCounts"
mmq = 30
if (sr_or_pe=="PE") {
  opt_pe="-p"
} else {
  opt_pe=" "  
}
args = paste0(
  " ", opt_pe, " -a ", saf_filename, " -F SAF -s 0 -Q ", mmq, " -T 32 -o ", prefix, "_count.txt ", 
  paste0(e$bam , " ", collapse=" "),
  ""
)
print(paste(cmd, args))
if (COUNT) {system2(cmd, args)}
```




##  Raw counts

```{r raw_counts, fig.height=3}
raw_counts_summary = read.table(paste0(prefix, "_count.txt.summary"), header=TRUE, row.names=1)
colnames(raw_counts_summary) = sub(paste0(".*", gse, "\\.([^.]+)_end\\.to\\.end.*"), "\\1", colnames(raw_counts_summary))
raw_counts_summary = raw_counts_summary[!apply(raw_counts_summary==0, 1, all),]
layout(matrix(1:3, 1), respect=TRUE)
cols = c(
  "#E63946",  # rouge
  "#F4A261",  # orange
  "#E9C46A",  # jaune
  "#2A9D8F",  # vert/bleu
  "#457B9D",  # bleu
  "#6D597A",  # violet
  "#B56576",  # vieux rose
  "#588157"   # vert
)
barplot(col=cols, as.matrix(raw_counts_summary), las=2)
barplot(col=cols, prop.table(as.matrix(raw_counts_summary), margin=2), las=2)
plot.new()
legend("topleft", legend=rownames(raw_counts_summary), fill=cols, bty="n")

raw_counts = read.table(paste0(prefix, "_count.txt"), header=TRUE)
rownames(raw_counts) = raw_counts[,1]
raw_counts = raw_counts[,-(1:6)]
colnames(raw_counts) = sub(paste0(".*", gse, "\\.([^.]+)_end\\.to\\.end.*"), "\\1", colnames(raw_counts))
head(raw_counts[,1:3])
tail(raw_counts[,1:3])
dim( raw_counts)
```

## Fingerprint

```{r fingerprint}
data = as.matrix(t(raw_counts))
col = as.numeric(e[,factoi[1]])
names(col) = rownames(e)

layout(1, respect=TRUE)
plot(0,0, col=0, xlim=c(1, ncol(data)), ylim=0:1, xlab="Ranked genomic features", ylab="Cumulative fraction of signal")
sapply(rownames(data), function(k){
  l = data[k,]
  tot = sum(sort(l))
  cs = cumsum(sort(l))
  lines(cs/tot, type="l", col=adjustcolor(col[k], alpha.f=0.7))
}) 
```


## CP100M norm counts

```{r norm_counts_cp100m, fig.height=4.5}
norm_counts_cp100m = t(t(raw_counts) / apply(raw_counts, 2, sum) * 100000000)
apply(norm_counts_cp100m, 2, sum)
layout(matrix(1:2, 1), respect=TRUE)
plot(density(norm_counts_cp100m))
plot(density(log2(norm_counts_cp100m+1)))

```

## DESeq2 norm counts


```{r norm_counts_deseq, fig.height=4.5}
if (!all(e$gsm %in% colnames(raw_counts))) {stop("problem gsm vs. colnames(raw_counts)")}
countData = raw_counts[,e$gsm]
colnames(countData) = rownames(e)
colData = e[colnames(countData), ]
# Vérifie que les noms de colonnes correspondent
if (!all(colnames(countData) == rownames(colData))) {stop("probleme data vs. e")}

dds = DESeq2::DESeqDataSetFromMatrix(
  countData = countData,
  colData = colData,
  design = ~ 1
)

dds = DESeq2::estimateSizeFactors(dds)

norm_counts_deseq = DESeq2::counts(dds, normalized=TRUE)
head(norm_counts_deseq[,1:3])
dim(norm_counts_deseq)

layout(matrix(1:2, 1), respect=TRUE)
plot(density(norm_counts_deseq))
plot(density(log2(norm_counts_deseq+1)))
```


# PCA


```{r pca_norm_counts}

norm_counts = list(
  DESeq2 = norm_counts_deseq, 
  CP100M = norm_counts_cp100m
)

for (k in names(norm_counts)) {
  print(k)
  data = log2(as.matrix(t(norm_counts[[k]]))+1)
  pca = prcomp(data, scale=FALSE)
  v = pca$sdev * pca$sdev
  p = v / sum(v) * 100
  # # --- MCLUST sur les scores PCA ---
  # library(mclust)
  # mc = mclust::Mclust(pca$x[,2:6], G=5)
  # cl = mc$classification
  # col = cl

  layout(matrix(1:6,2), respect=TRUE)
  # layout(matrix(1:2,1), respect=TRUE)
  barplot(p, ylab="% of variance explained", xlab="components")

  if (exists("factoi")) {
    col = as.numeric(e[rownames(pca$x),factoi[1]])
    legend("topright", col=1:length(levels(e[,factoi[1]])), levels(e[,factoi[1]]),pch=16)
  } else {
    col = 1
  }

  for (i in 1:5) {
    j = i+1
    plot(
      pca$x[,i], pca$x[,j],
      xlab=paste0("PC", i, "(", signif(p[i], 3), "%)"),
      ylab=paste0("PC", j, "(", signif(p[j], 3), "%)"),
      col=col,
      pch=16
    )
  }

  layout(1, respect=TRUE)

  i=1
  j = i+1
  plot(
    pca$x[,i], pca$x[,j],
    xlab=paste0("PC", i, "(", signif(p[i], 3), "%)"),
    ylab=paste0("PC", j, "(", signif(p[j], 3), "%)"),
    col=adjustcolor(col, alpha.f=0.5),
    pch=16
  )

  text(pca$x[,i], pca$x[,j], rownames(pca$x))
}
```





# Session Information

```{r results="verbatim"}
COUNT = FALSE
sessionInfo()
```



























