---
title: "DA on tiles"
author: "Florent Chuffart"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
---

```{r, echo=FALSE, eval=TRUE}
knitr::opts_chunk$set(collapse=TRUE, comment = "#>", fig.width=9, fig.height=6, eval=TRUE, echo=FALSE, results="hide")
```

# Pipeline Initialization

```{r}
source("common.R")
source("config")
prefix = "da_tiles"
```

# Samples

This section defines the experimental design and associates each sample with its corresponding sequencing files:

- `samples.R` contains the sample-specific information and defines the `samples` variable used throughout the pipeline.
- The `e` data frame defines the **experimental grouping**, with one row per sample and columns pointing to the corresponding BigWig (`bw`) and BAM (`bam`) files and factors if needed.
- The `gsm` column stores the GEO sample like identifiers, which are also used as row names to uniquely identify each sample.
- The file paths are constructed dynamically from the dataset (`gse`), sample name (`samples`), species, annotation, genome version, and sequencing configuration.

```{r label="samples"}
source("samples.R")
e = data.frame(
  gsm = samples,
  bam = paste0("/home/chuffarf/projects/datashare/", gse, "/", samples, "_end-to-end_trimno_bowtie2_", species, "_", annotation, "_", version, "_srt.bam"),
  bw = paste0("/home/chuffarf/projects/datashare/", gse, "/", samples, "_end-to-end_trim30_bowtie2_", species, "_", annotation, "_", version, "_srt_", sr_or_pe, "_30_4_RPKM.bw")
)
rownames(e) = e$gsm

if (file.exists("exp_grp.R")) {
  source("exp_grp.R")
}

head(e)
dim(e)
```



# DA

##  Raw counts

merge bam

macs3

peaks

```{r raw_counts_tiles}
raw_counts_summary = read.table(paste0("tiles_count.txt.summary"), header=TRUE, row.names=1)
colnames(raw_counts_summary) = sub(paste0(".*", gse, "\\.([^.]+)_end\\.to\\.end.*"), "\\1", colnames(raw_counts_summary))
layout(matrix(1:2, 1), respect=TRUE)
# barplot(table(as.matrix(raw_counts_summary), margin=2), las=2)
barplot(prop.table(as.matrix(raw_counts_summary), margin=2), las=2)

raw_counts = read.table(paste0("tiles_count.txt"), header=TRUE)
rownames(raw_counts) = raw_counts[,1]
pf = raw_counts[,1:6]
raw_counts = as.matrix(raw_counts[,-(1:6)])
colnames(raw_counts) = sub(paste0(".*", gse, "\\.([^.]+)_end\\.to\\.end.*"), "\\1", colnames(raw_counts))
head(raw_counts[,1:3])
tail(raw_counts[,1:3])
dim( raw_counts)
```


## Distribution

```{r density, results="verbatim"}
layout(matrix(1:2, 1), respect=TRUE)
plot(density(log2(raw_counts)+1))
# raw_counts = raw_counts[!apply(raw_counts<2048, 1, all),]
raw_counts = raw_counts[!apply(raw_counts<256, 1, all),]
plot(density(log2(raw_counts)+1))
dim(raw_counts)
```

## Deseq2 norm counts


```{r norm_counts_tiles, eval=FALSE}
if (!all(e$gsm %in% colnames(raw_counts))) {stop("problem gsm vs. colnames(raw_counts)")}
countData = raw_counts[,e$gsm]
colnames(countData) = rownames(e)
colData = e[colnames(countData), c("sprayed_status", "age", "DDEugg44", "self_spray")]
# Vérifie que les noms de colonnes correspondent
if (!all(colnames(countData) == rownames(colData))) {stop("probleme data vs. e")}


for (k in factoi[1]) {
  k = "sprayed_status"
  dds <- DESeq2::DESeqDataSetFromMatrix(
    countData = countData,
    colData = colData,
    # design = ~ age
    design = formula(paste("~", k))
    # design = ~ DDEugg44
  )
  dds <- DESeq2::DESeq(dds)
}

# plot(density(dds$DDEugg44))
# boxplot(dds$DDEugg44~dds$sprayed_status)


# barplot(table(dds$age))
# barplot(table(dds$age), las=2)
# barplot(table(dds$age, dds$sprayed_status), las=2)
# barplot(table(dds$sprayed_status, dds$age), las=2)


# boxplot(dds$sprayed_status~dds$DDEugg44)

# table(dds$sprayed_status)

# boxplot(dds$sprayed_status)
# boxplot(dds$sprayed_status)
# plot(dds$DDEugg44, dds$age)
# plot(dds$age, dds$DDEugg44)
# plot(density(dds$age))

# plot(dds$sprayed_status, dds$DDEugg44)
# plot(density(dds$DDEugg44))


# library(ggplot2)

# ggplot(e, aes(x = age, y = DDEugg44, colour = sprayed_status)) +
#   geom_point(size = 2, alpha = 0.7) +
#   geom_smooth(method = "lm", se = FALSE) +
#   theme_bw() +
#   labs(
#     x = "Age",
#     y = "DDEugg44",
#     colour = "Sprayed status"
#   )


# dds <- DESeq2::DESeq(dds)
res <- DESeq2::results(dds)

# dds = DESeq2::estimateSizeFactors(dds)
# norm_counts_tiles = DESeq2::counts(dds, normalized=TRUE)
# head(norm_counts_tiles[,1:3])
# dim(norm_counts_tiles)
```


# Volcano plots


```{r volcano, results="verbatim"}

if (!all(e$gsm %in% colnames(raw_counts))) {stop("problem gsm vs. colnames(raw_counts)")}
countData = raw_counts[,e$gsm]
colnames(countData) = rownames(e)
colData = e[colnames(countData), c("sprayed_status", "Smoking", "age", "DDEugg44", "self_spray")]
# Vérifie que les noms de colonnes correspondent
if (!all(colnames(countData) == rownames(colData))) {stop("probleme data vs. e")}

layout(matrix(1:2,1), respect=TRUE)

for (k in factoi[1:2]) {
  # k = "sprayed_status"
  print(k)
  dds <- DESeq2::DESeqDataSetFromMatrix(
    countData = countData,
    colData = colData,
    # design = ~ age
    design = formula(paste("~", k))
    # design = ~ DDEugg44
  )
  dds <- DESeq2::DESeq(dds)
  res <- DESeq2::results(dds)

  res = res[!is.na(res$pvalue),]
  res = res[!is.na(res$padj),]
  plot(res$log2FoldChange, -log10(res$pvalue))
  plot(res$log2FoldChange, -log10(res$padj))

  res = res[order(res$log2FoldChange), ]
  idx = rownames(res)[res$padj< 0.05]

  print(cbind(pf[idx,2:4], res[idx, c("log2FoldChange", "pvalue", "padj") ]))

}
```




# Session Information

```{r results="verbatim"}
sessionInfo()
```



























