---
title: "Fragment Size"
author: "Florent Chuffart"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
---


```{r}
prefix = "fragment_length"
max_frag_len = 500
options =   paste0(
    "--numberOfProcessors 32 ", 
    "--maxFragmentLength ", max_frag_len, " ", 
    "--binSize 10000 ", 
    "--distanceBetweenBins 1000000 "
  )
if (!exists("COMPUTEFRAGESIZE")) {COMPUTEFRAGESIZE = FALSE}
```

```{r, echo=FALSE, eval=TRUE}
knitr::opts_chunk$set(collapse=TRUE, comment = "#>", fig.width=9, fig.height=6, eval=TRUE, echo=FALSE, results="hide")
```

# Pipeline Initialization

This section initializes the environment required for the analysis:

- `common.R` contains the utility functions used throughout the pipeline.
- `config` contains the dataset-specific parameters used by the script, such as the project name, dataset name, and species.


```{r}
source("common.R")
source("config")
```

# Samples

This section defines the experimental design and associates each sample with its corresponding sequencing files:

- `samples.R` contains the sample-specific information and defines the `samples` variable used throughout the pipeline.
- The `e` data frame defines the **experimental grouping**, with one row per sample and columns pointing to the corresponding BigWig (`bw`) and BAM (`bam`) files and factors if needed.
- The `gsm` column stores the GEO sample like identifiers, which are also used as row names to uniquely identify each sample.
- The file paths are constructed dynamically from the dataset (`gse`), sample name (`samples`), species, annotation, genome version, and sequencing configuration.

```{r label="samples"}
source("samples.R")
e = data.frame(
  gsm = samples,
  bam = paste0("/home/chuffarf/projects/datashare/", gse, "/", samples, "_end-to-end_trimno_bowtie2_", species, "_", annotation, "_", version, "_srt.bam"),
  bw = paste0("/home/chuffarf/projects/datashare/", gse, "/", samples, "_end-to-end_trim30_bowtie2_", species, "_", annotation, "_", version, "_srt_", sr_or_pe, "_30_4_RPKM.bw")
)
rownames(e) = e$gsm
head(e)
dim(e)
```


# Fragment Size Distribution


```{r bamPEFragmentSize, eval=COMPUTEFRAGESIZE}
dir.create(prefix, showWarnings=FALSE)

foo = lapply(rownames(e), function(gsm) {
  bam_file = e[gsm,]$bam
  cmd = "bamPEFragmentSize"
  args = paste0(
    "-b ", bam_file, " ", 
    "-o ", prefix, "/", gsm, ".png ", 
    "--outRawFragmentLengths ", prefix, "/", gsm, ".txt ", 
    options
  )
  print(paste(cmd, args))
  system2(cmd, args)
})
```

## One histogram for all sample

```{r}
convs = sapply(rownames(e), function(gsm) {
  h = mread.table(paste0(prefix, "/", gsm, ".txt"), header=TRUE)
  conv = stats::ksmooth(h[,1], h[,2]/sum(h[,2]), "normal", bandwidth=5)
  conv
})

ylim = c(0, max(unlist(convs[2,]), na.rm=TRUE))
layout(1, respect=TRUE)
plot(0, 0, col=0, ylab="", xlab="", ylim=ylim, xlim=c(0,max_frag_len), main="")  
abline(v=c(75, 150, 300, 450), lty=3, col="grey")    
for (gsm in rownames(e)) {
  lines(convs[,gsm], col=adjustcolor(1, alpha.f=.5), lty=1, lwd=1)
}

```


## One histogram per sample

```{r hists}
layout(1, respect=TRUE)
for (gsm in rownames(e)) {
  print(gsm)
  plot(0, 0, col=0, ylab="", xlab="", ylim=ylim, xlim=c(0,max_frag_len), main=gsm)  
  abline(v=c(75, 150, 300, 450), lty=3, col="grey")    
  h = mread.table(paste0(prefix, "/", gsm, ".txt"), header=TRUE)
  lines(h[,1], h[,2]/sum(h[,2]), col=adjustcolor(2, alpha.f=.5), lty=1, lwd=1)
  conv = convs[,gsm]
  lines(conv, col=adjustcolor(1, alpha.f=.5), lty=1, lwd=1)
}
```


# PCA



```{r}
hs = lapply(rownames(e), function(gsm) {
  h = mread.table(paste0(prefix, "/", gsm, ".txt"), header=TRUE)
  h[,1]
})

tmp = sort(unique(unlist(hs)))
tmp = tmp[tmp<=max_frag_len]
sizes = paste0("sz", tmp)

data = matrix(0, nrow=length(rownames(e)), ncol=length(sizes))


rownames(data) = rownames(e)
colnames(data) = sizes
head(data[,1:10])

for (gsm in rownames(e)) {
  print(gsm)
  h = mread.table(paste0(prefix, "/", gsm, ".txt"), header=TRUE)
  h = h[h[,1] <= max_frag_len,]
  data[gsm, paste0("sz",h[,1])] = h[,2]  
}
head(data[,1:10])

layout(1, respect=TRUE)
image(t(data[rownames(pca$x)[order(pca$x[,1])], ]))

pca = prcomp(data, scale=FALSE)
v = pca$sdev * pca$sdev
p = v / sum(v) * 100
# # --- MCLUST sur les scores PCA ---
# library(mclust)
# mc = mclust::Mclust(pca$x[,2:6], G=5)
# cl = mc$classification
# col = cl

layout(matrix(1:6,2), respect=TRUE)
# layout(matrix(1:2,1), respect=TRUE)
barplot(p, ylab="% of variance explained", xlab="components")

if (file.exists("exp_grp.R")) {
  source("exp_grp.R")
  col = as.numeric(e[rownames(pca$x),factoi[1]])
  legend("topright", col=1:length(levels(e[,factoi[1]])), levels(e[,factoi[1]]),pch=16)
} else {
  col = 1
}

for (i in 1:5) {
  j = i+1
  plot(
    pca$x[,i], pca$x[,j],
    xlab=paste0("PC", i, "(", signif(p[i], 3), "%)"),
    ylab=paste0("PC", j, "(", signif(p[j], 3), "%)"),
    col=col,
    pch=16
  )
}

pca$x[rownames(pca$x)[order(pca$x[,1])],1]


layout(1, respect=TRUE)
image(t(data[rownames(pca$x)[order(pca$x[,1])], ]))


```


# Material and Methods

Fragment size distribution were optain using `bamPEFragmentSize` with options `r options`.




```{r results="verbatim"}
COMPUTEFRAGESIZE=FALSE
sessionInfo()
```



























