---
title: "Meta-Gene Analysis"
author: "Florent Chuffart"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
---

```{r, echo=FALSE, eval=TRUE}
knitr::opts_chunk$set(collapse=TRUE, comment = "#>", fig.width=9, fig.height=6, eval=TRUE, echo=FALSE, results="hide")
```

# Pipeline Initialization

This section initializes the environment required for the analysis:

- `common.R` contains the utility functions used throughout the pipeline.
- `config` contains the dataset-specific parameters used by the script, such as the project name, dataset name, and species.
- The `COUNTTILES` variable controls whether the read-counting step using `featureCounts` is performed and is set to `FALSE` by default.
- The `pca_tiles` prefix is used to reference the files generated by the pipeline.


```{r}
source("common.R")
source("config")
if (!exists("COUNTTILES")) {COUNTTILES = FALSE}
prefix = "metagene"
```

# Samples

This section defines the experimental design and associates each sample with its corresponding sequencing files:

- `samples.R` contains the sample-specific information and defines the `samples` variable used throughout the pipeline.
- The `e` data frame defines the **experimental grouping**, with one row per sample and columns pointing to the corresponding BigWig (`bw`) and BAM (`bam`) files and factors if needed.
- The `gsm` column stores the GEO sample like identifiers, which are also used as row names to uniquely identify each sample.
- The file paths are constructed dynamically from the dataset (`gse`), sample name (`samples`), species, annotation, genome version, and sequencing configuration.

```{r label="params"}
source("samples.R")
e = data.frame(
  gsm = samples,
  bam = paste0("/home/chuffarf/projects/datashare/", gse, "/", samples, "_end-to-end_trimno_bowtie2_", species, "_", annotation, "_", version, "_srt.bam"),
  bw = paste0("/home/chuffarf/projects/datashare/", gse, "/", samples, "_end-to-end_trimno_bowtie2_", species, "_", annotation, "_", version, "_srt_", sr_or_pe, "_30_4_RPKM.bw")
)
rownames(e) = e$gsm
head(e)
dim(e)
```


# Features


```{r label="tss_feature"}
regions_filenames = paste0("datashare/genomes/", species, "/", annotation, "/", version, "/Annotation/Genes/genes.bed")
regions_filenames
```

```{r results="verbatim"}
print(regions_filenames)
```




# Heatmap



```{r label="blacklist", eval=FALSE}
options(scipen=999)
if (species=="Mus_musculus" & version=="mm10") {
  # https://zenbu-wiki.gsc.riken.jp/zenbu/wiki/index.php/Uploading_UCSC_repetitive_elements_track
  # downloading UCSC rmsk data as BED
  # BED formatted UCSC track content can be obtained from UCSC table broswer.
  # The rmsk RepeatMasker (rmsk) track can be exported as BED file by selecting
  #
  # the assembly "Feb.2009 GRCh37/hg19"
  # the group "repeats and variations"
  # the track "RepeatMasker"
  # and finally the table "rmsk"
  # As we desire the complete repetitive elements genome-wide to be loaded into ZENBU, therefore we select
  #
  # region: "genome"
  # ZENBU enable gzip compressed bed files to be loaded directly, so we will further select :
  #
  # output format: "BED - browser extensibke format"
  # output file: we will name the file "UCSC_rmsk.hg19.bed.gz"
  # file type returned: "gzip compressed"

  rmsk = mread.table("~/projects/datashare/genomes/Mus_musculus/UCSC/mm10/rmsk.mm10.bed", skip=1, stringsAsFactors=FALSE)
  blacklist = rmsk[rmsk[,4] %in% c("GSAT_MM","SYNREP_MM"),]#, "(CTGTG)n"),]#
  blacklist = rbind(blacklist, list("chr19", 61266781, 61266782, "chr19:61266781-61266782", 1404, "+"))
  blacklist = rbind(blacklist, list("chr11", 3126500, 3200500, "Sfi1", 300, "-"))
  # sat2_feat_saf = rbind(sat2_feat_saf, list("Sfi1", "chr11", 3126500, 3200500, "-"))
  print(head(blacklist))
  dim(blacklist)
  blacklist_filename = "blacklist_file.bed"
  write.table(blacklist, file=blacklist_filename, sep="\t", quote=FALSE,row.names=FALSE, col.names=FALSE)      
} else if (species=="Homo_sapiens" & version=="hg38"){
  rmsk = mread.table("~/projects/datashare/genomes/Homo_sapiens/UCSC/hg38/rmsk.hg38.bed", skip=1, stringsAsFactors=FALSE)
  blacklist = rmsk[1,]
  blacklist = rbind(blacklist, list("chrM", 1, 20000, "chrM:1-20000", 20000, "+"))
  blacklist = blacklist[-1,]
  print(head(blacklist))
  dim(blacklist)
  blacklist_filename = "blacklist_file.bed"
  write.table(blacklist, file=blacklist_filename, sep="\t", quote=FALSE,row.names=FALSE, col.names=FALSE)      
} else if (species=="Candida_albicans" & version=="SC5314.A22"){
  blacklist_filename=NULL
} else if (species=="Candida_albicans" & version=="SC5314.A22.haplo"){
  blacklist_filename=NULL
} else if (species=="Saccharomyces_cerevisiae" & version=="sacCer3"){
  blacklist_filename=NULL
} else if (species=="Homo_sapiens" & version=="CHM13v2.0"){
  blacklist_filename=NULL
} else {
  stop("genome not Homo_sapiens hg38 nor Mus_musculus mm10 nor Candida_albicans SC5314.A22 nor Saccharomyces_cerevisiae sacCer3")
}
```

# Compute matrix

```{r, label="computeMatrix tss_keys", eval=TRUE}
# stop("EFN")
dir.create("tmp", showWarnings=FALSE)
matrix_filename = paste0("tmp/matrix_", prefix, ".txt.gz")
# score_filenames
idx_samples = 1:2
score_filenames = e$bw[idx_samples]
names(score_filenames) = rownames(e)[idx_samples]
brsl = arsl = 5000
bin_size = 100

cmd = "computeMatrix"
args = paste0("reference-point ", 
    "-S ", paste0(score_filenames, collapse=" "), " ",  
    "-R ", paste0(regions_filenames, collapse=" "), " ", 
    "--outFileName ", matrix_filename, " ", 
    "--referencePoint TSS ", 
    "--binSize ", bin_size, " ", 
    "--beforeRegionStartLength ", brsl, " ", 
    "--afterRegionStartLength ", arsl, " ", 
    "--numberOfProcessors 32 ", 
    "--sortRegions descend ", 
    # "--blackListFileName blacklist_file.bed", 
    NULL
  )
print(paste(cmd, args))
system2(cmd, args)

```

# Heatmap



```{r heatmap, eval=TRUE}
hm_out_filename = paste0("tmp/hm_metagene_", prefix, ".png")
cmd = "plotHeatmap"
args = paste0(
    "--matrixFile ", matrix_filename, " ", 
    "--outFileName ", hm_out_filename, " ", 
    "--colorMap YlOrRd ", 
    "--sortRegions descend ", 
    "--samplesLabel ", paste0(names(score_filenames), collapse=" "), " ", 
    # "--kmeans 4 ", 
    # "--outFileSortedRegions feat_kmeans4_", prefix, ".bed.csv ", 
    NULL
  )
print(paste(cmd, args))
system2(cmd, args)
```

```{r, echo=FALSE, eval=TRUE, out.width="100%", results="markup"}
knitr::include_graphics(hm_out_filename)
```




# Profile


```{r profile}
pf_out_filename = paste0("tmp/pf_metagene_", prefix, ".png")
cmd = "plotProfile"
args = paste0(
    "--matrixFile tmp/matrix_", prefix, ".txt.gz ", 
    "--outFileName ", pf_out_filename, " ", 
    "--samplesLabel ", paste0(names(score_filenames), collapse=" "), " ", 
    "--perGroup ", 
    # " --kmeans 4", 
    NULL
  )
print(paste(cmd, args))
system2(cmd, args)
```



```{r, echo=FALSE, out.width="100%", results="markup"}
knitr::include_graphics(pf_out_filename)
```






```{r, eval=TRUE}
# dir.create("tmp", showWarnings=FALSE)

# matrix_file = paste0("tmp/matrix_", prefix, ".txt.gz")

# # score_filenames
# score_filenames = paste0("datashare/", gse, "/", samples[1], "_end-to-end_trim30_bowtie2_", species, "_", annotation, "_", version, "_srt_", sr_or_pe, "_30_4_RPKM.bw")
# brsl = arsl = 5000
# bin_size = 100
# Sys.setenv(PATH="/summer/epistorage/miniconda3/envs/mnase_env/bin:/summer/epistorage/miniconda3/condabin:/home/chuffarf/perl5/bin:/summer/epistorage/opt/bin:/usr/local/bin:/usr/bin:/bin:/usr/local/games:/usr/games")
# matrix_file = deeptoolsr::dt_compute_matrix(regions_filename=regions_filenames, score_filename=score_filenames, out_filename=matrix_file,
#   blacklist_filename = blacklist_filename,
#   bin_size = bin_size,
#   before_region_start_length = brsl,
#   after_region_start_length = arsl,
#   FORCE_EXEC=TRUE,
#   number_of_processors=32,
#   sort_regions="descend"
# )
# ```


# ```{r eval=TRUE, label="sort_heatmap"}
# source("common.R")
# ordered_wig_signal = fast_matrix_file(matrix_file, GRP=TRUE)
# hm_out_filename = paste0("tmp/hm_", prefix, ".fasted.png")
# hm_out_filename = deeptoolsr::dt_plot_heatmap(matrix_file=paste0(matrix_file, ".fast.txt.gz"), out_filename=hm_out_filename, FORCE_EXEC=TRUE, color_map="YlOrRd", sort_regions="descend")
# ```

# ```{r, echo=FALSE, eval=TRUE, out.width="100%", results="markup"}
# knitr::include_graphics(hm_out_filename)
# ```



# ```{r, eval=FALSE}
# raw_wig_signal = read.table(gzfile(matrix_file), skip=1, stringsAsFactors=FALSE, na.strings="nan")
# foo = apply(raw_wig_signal[,-(1:6)], 1, max)
# rev(sort(foo))[1:6]
# head(raw_wig_signal[rev(rank(foo)),])
# table(raw_wig_signal[,1])
```




# Session Information

```{r results="verbatim"}
sessionInfo()
```



























