---
title: "PCA on tiles"
author: "Florent Chuffart"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
---

```{r, echo=FALSE, eval=TRUE}
knitr::opts_chunk$set(collapse=TRUE, comment = "#>", fig.width=9, fig.height=6, eval=TRUE, echo=FALSE, results="hide")
```

# Pipeline Initialization

This section initializes the environment required for the analysis:

- `common.R` contains the utility functions used throughout the pipeline.
- `config` contains the dataset-specific parameters used by the script, such as the project name, dataset name, and species.
- The `COUNTTILES` variable controls whether the read-counting step using `featureCounts` is performed and is set to `FALSE` by default.
- The `pca_tiles` prefix is used to reference the files generated by the pipeline.


```{r}
source("common.R")
source("config")
if (!exists("COUNTTILES")) {COUNTTILES = FALSE}
prefix = "pca_tiles"
```

# Samples

This section defines the experimental design and associates each sample with its corresponding sequencing files:

- `samples.R` contains the sample-specific information and defines the `samples` variable used throughout the pipeline.
- The `e` data frame defines the **experimental grouping**, with one row per sample and columns pointing to the corresponding BigWig (`bw`) and BAM (`bam`) files and factors if needed.
- The `gsm` column stores the GEO sample like identifiers, which are also used as row names to uniquely identify each sample.
- The file paths are constructed dynamically from the dataset (`gse`), sample name (`samples`), species, annotation, genome version, and sequencing configuration.

```{r label="samples"}
source("samples.R")
e = data.frame(
  gsm = samples,
  bam = paste0("/home/chuffarf/projects/datashare/", gse, "/", samples, "_end-to-end_trimno_bowtie2_", species, "_", annotation, "_", version, "_srt.bam"),
  bw = paste0("/home/chuffarf/projects/datashare/", gse, "/", samples, "_end-to-end_trim30_bowtie2_", species, "_", annotation, "_", version, "_srt_", sr_or_pe, "_30_4_RPKM.bw")
)
rownames(e) = e$gsm
head(e)
dim(e)
```



# Genome

This section defines the genomic coordinates used throughout the analysis.

```{r genome}
# to create .chrom.sizes file for Mus_musculus/UCSC/mm10 do 
# cat ~/projects/datashare/genomes/Mus_musculus/UCSC/mm10/Sequence/WholeGenomeFasta/genome.dict | grep SN | cut -f2,3 | cut -d: -f2,3 | sed -e "s/LN://" > ~/projects/datashare/genomes/Mus_musculus/UCSC/mm10/mm10.chrom.sizes
chr_sizes = read.table(paste0("~/projects/datashare/genomes/", species, "/", annotation, "/", version, "/", version, ".chrom.sizes"), stringsAsFactors=FALSE)
head(chr_sizes)
dim(chr_sizes)
# Filtering...
# chr_sizes = chr_sizes[chr_sizes[,1] != "chrM",]
# chr_sizes = chr_sizes[c(11:19, 1:10, 20:21),]
chr_sizes = chr_sizes[1:24,]
genome = data.frame(chr=chr_sizes[,1], beg=1, end=chr_sizes[,2], name1=chr_sizes[,1], name2=chr_sizes[,1], stringsAsFactorsnd="+", stringsAsFactors=FALSE)

rownames(genome) = NULL
print(genome)
dim(genome)
```

```{r, results="verbatim", echo=TRUE}
genome
```

# Features

This section defines the genomic features used for the downstream analysis.

```{r results="verbatim"}
feat = tile_genome(5000, genome)
head(feat)
dim(feat)
```


## Counts


```{r count_tiles, eval=COUNTTILES}
# Counts
feat_saf = feat[, c(4,1:3, 6)]
# feat_saf = feat[1:100,c(4,1:3, 6)]
colnames(feat_saf) = c("GeneID", "Chr", "Start", "End", "Strand")
feat_saf$Start = as.numeric(feat_saf$Start)
feat_saf$End   = as.numeric(feat_saf$End  )
saf_filename = paste0(prefix, ".saf")
write.table(feat_saf, file=saf_filename, sep="\t", quote=FALSE,row.names=FALSE, col.names=FALSE)


cmd = "/summer/epistorage/miniconda3/envs/subread_env/bin/featureCounts"
cmd = "featureCounts"
mmq = 30
# if PE
if (sr_or_pe=="PE") {
  opt_pe="-p"
} else {
  opt_pe=" "  
}
args = paste0(
  " ", opt_pe, " -a ", saf_filename, " -F SAF -s 0 -Q ", mmq, " -T 32 -o ", prefix, "_count.txt ", 
  paste0(e$bam , " ", collapse=" "),
  ""
)
print(paste(cmd, args))
system2(cmd, args)
system2("cat", paste0(prefix, "_count_.txt.summary"))
# stop("EFN")
```




##  Raw counts

```{r raw_counts_tiles}
raw_counts = read.table(paste0(prefix, "_count.txt"), header=TRUE)
rownames(raw_counts) = raw_counts[,1]
raw_counts = raw_counts[,-(1:6)]
colnames(raw_counts) = sub(paste0(".*", gse, "\\.([^.]+)_end\\.to\\.end.*"), "\\1", colnames(raw_counts))
head(raw_counts[,1:3])
tail(raw_counts[,1:3])
dim( raw_counts)
```

## CP50M norm counts

```{r}
```

## Deseq2 norm counts


```{r norm_counts_tiles, eval=TRUE}
if (!all(e$gsm %in% colnames(raw_counts))) {stop("problem gsm vs. colnames(raw_counts)")}
countData = raw_counts[,e$gsm]
colnames(countData) = rownames(e)
colData = e[colnames(countData), ]
# Vérifie que les noms de colonnes correspondent
if (!all(colnames(countData) == rownames(colData))) {stop("probleme data vs. e")}

dds <- DESeq2::DESeqDataSetFromMatrix(
  countData = countData,
  colData = colData,
  design = ~ 1
)

# dds <- DESeq2::DESeq(dds)
# res_age <- DESeq2::results(dds, name = "age")

dds = DESeq2::estimateSizeFactors(dds)

norm_counts_tiles = DESeq2::counts(dds, normalized=TRUE)
head(norm_counts_tiles[,1:3])
dim(norm_counts_tiles)
```


# PCA


```{r pca_norm_counts_tiles}
data = as.matrix(t(norm_counts_tiles))
pca = prcomp(data, scale=FALSE)
v = pca$sdev * pca$sdev
p = v / sum(v) * 100
# # --- MCLUST sur les scores PCA ---
# library(mclust)
# mc = mclust::Mclust(pca$x[,2:6], G=5)
# cl = mc$classification
# col = cl

layout(matrix(1:6,2), respect=TRUE)
# layout(matrix(1:2,1), respect=TRUE)
barplot(p, ylab="% of variance explained", xlab="components")

if (file.exists("exp_grp.R")) {
  source("exp_grp.R")
  col = as.numeric(e[rownames(pca$x),factoi[1]])
  legend("topright", col=1:length(levels(e[,factoi[1]])), levels(e[,factoi[1]]),pch=16)
} else {
  col = 1
}

for (i in 1:5) {
  j = i+1
  plot(
    pca$x[,i], pca$x[,j],
    xlab=paste0("PC", i, "(", signif(p[i], 3), "%)"),
    ylab=paste0("PC", j, "(", signif(p[j], 3), "%)"),
    col=col,
    pch=16
  )
}

layout(1, respect=TRUE)

i=1
j = i+1
plot(
  pca$x[,i], pca$x[,j],
  xlab=paste0("PC", i, "(", signif(p[i], 3), "%)"),
  ylab=paste0("PC", j, "(", signif(p[j], 3), "%)"),
  col=adjustcolor(col, alpha.f=0.5),
  pch=16
)

text(pca$x[,i], pca$x[,j], rownames(pca$x))

```






# Session Information

```{r results="verbatim"}
COUNTTILES = FALSE
sessionInfo()
```



























