---
title: "Count features with `featureCounts`, and export count files for DESeq2"
author: "Florent Chuffart"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
---


```{r, echo=FALSE, eval=TRUE}
knitr::opts_chunk$set(collapse=TRUE, comment = "#>", fig.width=9, fig.height=6, eval=TRUE, echo=FALSE, results="hide")
source("config")
source("config.R")
source("common.R")
```

```{r params}
gsms = samples[c(3,5,7,9,13,15,17, 19)]
locus_start = -300
locus_end = 0
mmq = 30
tmp_prefix = paste0("featureCounts_tss_", locus_start, ":", locus_end, "_mmq", mmq)
```

# Features

In this section, genomic features are defined around transcription start sites (TSS) to quantify sequencing signal in promoter-proximal regions. Gene coordinates are imported from the reference genome annotation and converted into strand-aware intervals spanning the TSS (`r locus_start`:`r locus_end`). The resulting regions are formatted into a SAF (Simplified Annotation Format) annotation file, which can subsequently be used for read counting and downstream analyses. 

```{r}
# get genes as bed
genes_file = paste0("~/projects/datashare/genomes/", species, "/", annotation, "/", version, "/Annotation/Genes/genes.bed")
genes_orig = genes = read.table(genes_file, header=FALSE)

# build features as bed
tss = genes
tss[,2] = ifelse(genes$V6=="+", genes$V2 + locus_start, genes$V3 - locus_end)
tss[tss[,2]<1,2] = 1 # when tss is so close oh the beginning of the chromosome
tss[,3] = tss[,2] + (locus_end - locus_start)

# convert to saf
feat_saf = tss[, c(4,1:3, 6)]
colnames(feat_saf) = c("GeneID", "Chr", "Start", "End", "Strand")
saf_filename = paste0(tmp_prefix, ".saf")
write.table(feat_saf, file=saf_filename, sep="\t", quote=FALSE,row.names=FALSE, col.names=FALSE)
```

# Counts

Read quantification was performed using `featureCounts` on paired-end aligned BAM files. Sequencing reads were assigned to the previously defined promoter-associated TSS regions provided in SAF format. Only reads with a mapping quality score greater than or equal to `r mmq` were retained. Counting was executed in unstranded mode using 16 threads to improve computational efficiency, and summary statistics were generated to assess assignment rates across samples.

```{r}
cmd = "/summer/epistorage/miniconda3/envs/subread_env/bin/featureCounts"
cmd = "featureCounts"
rawcounts_file = paste0(tmp_prefix, ".txt")
summary_file = paste0(rawcounts_file, ".summary")
bam_files = paste0("/home/chuffarf/projects/datashare/", gse, "/", gsms, "_end-to-end_trim30_bowtie2_", species, "_", annotation, "_", version, "_srt.bam")
args = paste0(
  "-p -a ", saf_filename, " -F SAF -s 0 -Q ", mmq, " -T 16 -o ", rawcounts_file, " ", 
  paste0(bam_files, collapse=" "),
  ""
)
print(paste(cmd, args))
system2(cmd, args)

summary = read.table(summary_file, header=TRUE)
raw_counts = read.table(rawcounts_file, header=TRUE)
colnames(summary)[-1] = colnames(raw_counts)[7:ncol(raw_counts)] = (tmp = do.call(rbind, strsplit(bam_files, "/|_end")))[,ncol(tmp)-1]

```{r}
# Convert to matrix for plotting
counts_mat_summary = as.matrix(summary[, -1])
rownames(counts_mat_summary) = summary$Status
# Remove empty rows
counts_mat_summary = counts_mat_summary[!apply(counts_mat_summary==0, 1, all),]
# Barplot
layout(1, respect=TRUE)
par(mar=c(15, 4.1, 4.1, 2.1))
barplot(
  counts_mat_summary,
  beside = FALSE,
  col = rainbow(nrow(counts_mat_summary)),
  las = 2,
  cex.names = 0.8,
  main = "featureCounts assignment summary",
  ylab = "Number of reads"
)
# Legend
legend("topright", rownames(counts_mat_summary), fill=rainbow(nrow(counts_mat_summary)))
par(mar=c(5.1, 4.1, 4.1, 2.1))
```
# Export

```{r}
raw_counts[,7:ncol(raw_counts)]

for i in 1:length(bam_filess) {
  write.table(raw_counts[,c(1, 6+i)], paste0(bam_files[i], "_", tmp_prefix, ".txt"), sep="\t", quote=FALSE, row.names=FALSE, col.names=FALSE)
}
```

# Session Information

```{r results="verbatim"}
sessionInfo()
date()
```



























