---
title: "Merging bam files"
author: "Florent Chuffart"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
---




```{r}
prefix = "merge_bam"
if (!exists("MERGEBAM")) {MERGEBAM = FALSE}
```

```{r, echo=FALSE, eval=TRUE}
knitr::opts_chunk$set(collapse=TRUE, comment = "#>", fig.width=9, fig.height=6, eval=TRUE, echo=FALSE, results="hide")
```

# Pipeline Initialization

This section initializes the environment required for the analysis:

- `common.R` contains the utility functions used throughout the pipeline.
- `config` contains the dataset-specific parameters used by the script, such as the project name, dataset name, and species.


```{r}
source("common.R")
source("config")
```

# Samples

This section defines the experimental design and associates each sample with its corresponding sequencing files:

- `samples.R` contains the sample-specific information and defines the `samples` variable used throughout the pipeline.
- The `e` data frame defines the **experimental grouping**, with one row per sample and columns pointing to the corresponding BigWig (`bw`) and BAM (`bam`) files and factors if needed.
- The `gsm` column stores the GEO sample like identifiers, which are also used as row names to uniquely identify each sample.
- The file paths are constructed dynamically from the dataset (`gse`), sample name (`samples`), species, annotation, genome version, and sequencing configuration.

```{r label="samples"}
source("samples.R")
e = data.frame(
  gsm = samples,
  bam = paste0("/home/chuffarf/projects/datashare/", gse, "/", samples, "_end-to-end_trimno_bowtie2_", species, "_", annotation, "_", version, "_srt.bam"),
  bw = paste0("/home/chuffarf/projects/datashare/", gse, "/", samples, "_end-to-end_trimno_bowtie2_", species, "_", annotation, "_", version, "_srt_", sr_or_pe, "_30_4_RPKM.bw")
)
rownames(e) = e$gsm

if (file.exists("exp_grp.R")) {
  source("exp_grp.R")
}

head(e)
dim(e)
# stop("EFN")
```

# Merge bams


```{r mergebam, eval=TRUE, echo=TRUE, results="verbatim"}
idx = rownames(e)
bamgrpprefix = ""

idx = 1:3
bamgrpprefix = "123"

idx = rownames(e)[e$sprayed_status=="Sprayed"]
bamgrpprefix = "Sprayed"

idx = rownames(e)[e$sprayed_status=="Unsprayed"]
bamgrpprefix = "Unsprayed"


temp_dir = list.files("/localscratch/", pattern="chuffarf*", full.names=TRUE)

# merge all bams
cmd = "samtools"
args = paste0("merge ", 
    "-@ 32 ", 
    # "-T ", temp_dir, " ", 
    "", temp_dir, "/merged", bamgrpprefix, "_end-to-end_trimno_bowtie2_", species, "_", annotation, "_", version, ".bam ", 
    paste0("datashare/", gse, "/", e[idx,]$gsm, "_end-to-end_trimno_bowtie2_", species, "_", annotation, "_", version, "_srt.bam", collapse=" "), 
    NULL
  )
print(paste(cmd, args))
system2(cmd, args)
print("done.")

# filter merged bam file
cmd = "samtools"
args = paste0("view ", 
    "-@ 32 ", 
    "-b -f 2 -q 30 ", 
    "-e 'tlen > -500 && tlen < 500 && tlen != 0' ", 
    "-o ", temp_dir, "/merged", bamgrpprefix, "_end-to-end_trimno_bowtie2_", species, "_", annotation, "_", version, "_fltr.bam ",
    "", temp_dir, "/merged", bamgrpprefix, "_end-to-end_trimno_bowtie2_", species, "_", annotation, "_", version, ".bam ", 
    NULL
  )
print(paste(cmd, args))
system2(cmd, args)
print("done.")

# rsync filtered bam to datashare
cmd = "rsync"
args = paste0( "-auvP ", 
    "", temp_dir,      "/merged", bamgrpprefix, "_end-to-end_trimno_bowtie2_", species, "_", annotation, "_", version, "_fltr.bam ",
    "datashare/", gse, "/merged", bamgrpprefix, "_end-to-end_trimno_bowtie2_", species, "_", annotation, "_", version, "_fltr.bam ",
    NULL
  )
print(paste(cmd, args))
system2(cmd, args)
print("done.")

# clean temp_dir
cmd = "rm"
args = paste0( "-Rf ", 
    "", temp_dir, "/*", 
    NULL
  )
print(paste(cmd, args))
system2(cmd, args)
print("done.")

# sort filtered bam
cmd = "samtools"
args = paste0("sort ", 
    "-@ 32 ", 
    "-m 2G ", 
    "-T ", temp_dir, "/sort_tmp ", 
    "datashare/", gse, "/merged", bamgrpprefix, "_end-to-end_trimno_bowtie2_", species, "_", annotation, "_", version, "_fltr.bam ",
    "-o ", temp_dir,   "/merged", bamgrpprefix, "_end-to-end_trimno_bowtie2_", species, "_", annotation, "_", version, "_fltr_srt.bam ", 
    NULL
  )
print(paste(cmd, args))
system2(cmd, args)
print("done.")

# rsync filtered sorted bam to datashare
cmd = "rsync"
args = paste0( "-auvP ", 
    "", temp_dir,      "/merged", bamgrpprefix, "_end-to-end_trimno_bowtie2_", species, "_", annotation, "_", version, "_fltr_srt.bam ",
    "datashare/", gse, "/merged", bamgrpprefix, "_end-to-end_trimno_bowtie2_", species, "_", annotation, "_", version, "_fltr_srt.bam ",
    NULL
  )
print(paste(cmd, args))
system2(cmd, args)
print("done.")

# index it
cmd = "samtools"
args = paste0("index ", 
    "-@ 32 ", 
    "datashare/", gse, "/merged", bamgrpprefix, "_end-to-end_trimno_bowtie2_", species, "_", annotation, "_", version, "_fltr_srt.bam ",
    NULL
  )
print(paste(cmd, args))
system2(cmd, args)
print("done.")

# rm merged filtered bam
cmd = "rm"
args = paste0( 
    "datashare/", gse, "/merged", bamgrpprefix, "_end-to-end_trimno_bowtie2_", species, "_", annotation, "_", version, "_fltr.bam ",
    NULL
  )
print(paste(cmd, args))
system2(cmd, args)
print("done.")

# clean temp_dir
cmd = "rm"
args = paste0( "-Rf ", 
    "", temp_dir, "/*", 
    NULL
  )
print(paste(cmd, args))
system2(cmd, args)
print("done.")

# bigwig
bin_size = 4
mmq = 30
cmd = "bamCoverage"
norm = "RPKM"
args = paste0(
    "--numberOfProcessors 32 ", 
    "--extendReads ", 
    "--centerReads ", 
    "--normalizeUsing ", norm, " ", 
    "--binSize ", bin_size, " ", 
    "--minMappingQuality ", mmq, " ", 
    " -b datashare/", gse, "/merged", bamgrpprefix, "_end-to-end_trimno_bowtie2_", species, "_", annotation, "_", version, "_fltr_srt.bam ", 
    " -o datashare/", gse, "/merged", bamgrpprefix, "_end-to-end_trimno_bowtie2_", species, "_", annotation, "_", version, "_fltr_srtcr_PE_", mmq, "_", bin_size, "_", norm, ".bw ",
    NULL
  )
print(paste(cmd, args))
system2(cmd, args)

```






# Session Information

```{r results="verbatim"}
sessionInfo()
```



