---
title: "Merging bam files"
author: "Florent Chuffart"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
---




```{r}
prefix = "merge_bam"
if (!exists("MERGEBAM")) {MERGEBAM = FALSE}
```

```{r, echo=FALSE, eval=TRUE}
knitr::opts_chunk$set(collapse=TRUE, comment = "#>", fig.width=9, fig.height=6, eval=TRUE, echo=FALSE, results="hide")
```

# Pipeline Initialization

This section initializes the environment required for the analysis:

- `common.R` contains the utility functions used throughout the pipeline.
- `config` contains the dataset-specific parameters used by the script, such as the project name, dataset name, and species.


```{r}
source("common.R")
source("config")
```

# Samples

This section defines the experimental design and associates each sample with its corresponding sequencing files:

- `samples.R` contains the sample-specific information and defines the `samples` variable used throughout the pipeline.
- The `e` data frame defines the **experimental grouping**, with one row per sample and columns pointing to the corresponding BigWig (`bw`) and BAM (`bam`) files and factors if needed.
- The `gsm` column stores the GEO sample like identifiers, which are also used as row names to uniquely identify each sample.
- The file paths are constructed dynamically from the dataset (`gse`), sample name (`samples`), species, annotation, genome version, and sequencing configuration.

```{r label="samples"}
source("samples.R")
e = data.frame(
  gsm = samples,
  bam = paste0("/home/chuffarf/projects/datashare/", gse, "/", samples, "_end-to-end_trimno_bowtie2_", species, "_", annotation, "_", version, "_srt.bam"),
  bw = paste0("/home/chuffarf/projects/datashare/", gse, "/", samples, "_end-to-end_trimno_bowtie2_", species, "_", annotation, "_", version, "_srt_", sr_or_pe, "_30_4_RPKM.bw")
)
rownames(e) = e$gsm
head(e)
dim(e)
stop("EFN")
```

# ATAC peaks



```{r mergebammacs3, eval=FALSE}
# locus = "chr5:155000000-165000000"
# cmd = "samtools"
# args = paste0("merge - ", paste0("  <(samtools view -bh /home/chuffarf/projects/datashare/", gse, "/", idx_samples, "_end-to-end_trimno_bowtie2_", species, "_", annotation, "_", version, "_srt_mmq30.bam ", locus, " ) ", collapse=" "), " > /home/chuffarf/projects/datashare/", gse, "/merged", locus, "_end-to-end_trimno_bowtie2_", species, "_", annotation, "_", version, "_srt_mmq30.bam")
# print(paste(cmd, args))
# writeLines(paste(cmd, args), "tmp_mergebam.sh")
# system2("bash", "tmp_mergebam.sh")

# merge
cmd = "samtools"
args = paste0("merge ", 
    "-@ 32 ", 
    "datashare/", gse, "/merged_end-to-end_trimno_bowtie2_", species, "_", annotation, "_", version, ".bam ", 
    paste0("datashare/", gse, "/", e$gsm, "_end-to-end_trimno_bowtie2_", species, "_", annotation, "_", version, "_srt.bam", collapse=" "), 
    NULL
  )
print(paste(cmd, args))
system2(cmd, args)

# sort
cmd = "samtools"
args = paste0("sort ", 
    "-@ 32 ", 
    "-m 2G ", 
    "datashare/", gse, "/merged_end-to-end_trimno_bowtie2_", species, "_", annotation, "_", version, "_srt_mmq30.bam -o /home/chuffarf/projects/datashare/", gse, "/merged_end-to-end_trimno_bowtie2_", species, "_", annotation, "_", version, "_srt.bam ", 
    NULL
  )
print(paste(cmd, args))
system2(cmd, args)

# index
cmd = "samtools"
args = paste0("index  -@ 32   /home/chuffarf/projects/datashare/", gse, "/merged_end-to-end_trimno_bowtie2_", species, "_", annotation, "_", version, "_srt.bam ")
print(paste(cmd, args))
system2(cmd, args)

# bigwig
bin_size = 4
mmq = 30
cmd = "bamCoverage"
norm = "RPKM"
args = paste0(
    "--numberOfProcessors 32 ", 
    "--extendReads ", 
    "--normalizeUsing ", norm, " ", 
    "--binSize ", bin_size, " ", 
    "--minMappingQuality ", mmq, " ", 
    " -b datashare/", gse, "/merged_end-to-end_trimno_bowtie2_", species, "_", annotation, "_", version, "_srt.bam ", 
    " -o datashare/", gse, "/merged_end-to-end_trimno_bowtie2_", species, "_", annotation, "_", version, "_srt_PE_", mmq, "_", bin_size, "_", norm, ".bw "
  )
print(paste(cmd, args))
system2(cmd, args)

# bigwig
bin_size = 4
mmq = 30
cmd = "bamCoverage"
norm = "RPKM"
args = paste0(
    "--numberOfProcessors 32 ", 
    "--extendReads ", 
    "--centerReads ", 
    "--normalizeUsing ", norm, " ", 
    "--binSize ", bin_size, " ", 
    "--minMappingQuality ", mmq, " ", 
    " -b datashare/", gse, "/merged_end-to-end_trimno_bowtie2_", species, "_", annotation, "_", version, "_srt.bam ", 
    " -o datashare/", gse, "/merged_end-to-end_trimno_bowtie2_", species, "_", annotation, "_", version, "_srtcr_PE_", mmq, "_", bin_size, "_", norm, ".bw "
  )
print(paste(cmd, args))
system2(cmd, args)


for (tmp_thresh in c("1.00e-25", "1.00e-50", "1.00e-99")[3]) {
  cmd = "macs3"
  args = paste0("callpeak --call-summits -n mrg", tmp_thresh, " -t /home/chuffarf/projects/datashare/", gse, "/merged_end-to-end_trimno_bowtie2_", species, "_", annotation, "_", version, "_srt_mmq30_srt.bam -f BAMPE -g hs --keep-dup all -q ", tmp_thresh, " --outdir macs3_output_merged ")
  print(paste(cmd, args))
  system2(cmd, args)
}

```

## Stats

```{r peaks, fig.height=6}
# #	Nom de la colonne	Description	Type / Exemple
# 1	chrom	Nom du chromosome	chr1
# 2	chromStart	Position de début du pic (0-based, inclusif)	3455000
# 3	chromEnd	Position de fin du pic (1-based, exclusif)	3455500
# 4	name	Nom du pic	peak_00001
# 5	score	Score de 0 à 1000 (utile pour visualisation UCSC)	500
# 6	strand	Brin (+, -, ou . si non applicable)	.
# 7	signalValue	Enrichissement du signal au sommet du pic	12.45
# 8	pValue	–log10(p-value) de la significativité du pic	7.30
# 9	qValue	–log10(q-value, FDR ajustée)	5.21
# 10	peak	Position du sommet du pic par rapport à chromStart (en bp). -1 si inconnu.	250
# peak_prefix = "mrg1.00e-25_peaks"
# peak_prefix = "mrg1.00e-50_peaks"
peak_prefix = "mrg1.00e-99_peaks"
peaks = read.table(paste0("macs3_output_merged/", peak_prefix, ".narrowPeak"))
rownames(peaks) = peaks[,4]
colnames(peaks) = c("chrom", "chromStart", "chromEnd", "name", "score", "strand", "signalValue", "pValue", "qValue", "peak")
peaks$len = peaks$chromEnd - peaks$chromStart
peaks = peaks[nchar(peaks$chrom)<=5,]
head(peaks)
dim(peaks)

# Filtering peaks
head(peaks)
dim(peaks)
peaks$key = paste0(peaks[[1]], ":", peaks[[2]], "-", peaks[[3]])
length(unique(peaks$key))
qv_thresh = peaks[order(peaks$qValue, decreasing=TRUE),]$qValue[40000]
peaks = peaks[peaks$qValue>=qv_thresh,]
dim(peaks)
length(unique(peaks$key))
peaks = peaks[!duplicated(peaks$key),]
dim(peaks)
length(unique(peaks$key))

# Stat peaks
custom_hm(peaks$qValue , peaks$pValue     , xlab="qValue", ylab="pValue     ", LOGCOUNT=TRUE, main=peak_prefix)#, xlim=c(0, quantile(peaks$qValue, .99), ylim=c(0, quantile(peaks$pValue     , .99))))
custom_hm(peaks$qValue , peaks$signalValue, xlab="qValue", ylab="signalValue", LOGCOUNT=TRUE, main=peak_prefix)#, xlim=c(0, quantile(peaks$qValue, .99), ylim=c(0, quantile(peaks$signalValue, .99))))
custom_hm(peaks$len    , peaks$signalValue, xlab="len   ", ylab="signalValue", LOGCOUNT=TRUE, main=peak_prefix)#, xlim=c(0, quantile(peaks$len   , .99), ylim=c(0, quantile(peaks$signalValue, .99))))
custom_hm(peaks$len    , peaks$qValue     , xlab="len   ", ylab="qValue     ", LOGCOUNT=TRUE, main=peak_prefix)#, xlim=c(0, quantile(peaks$len   , .99), ylim=c(0, quantile(peaks$qValue     , .99))))
# stop("EFN")
```

```{r prefix2}
prefix2 = paste0(prefix, "_", peak_prefix)
```

## TFBS


```{r call_jaspar fun}
library(TFBSTools)
library(JASPAR2022)
library(Biostrings)

if (!exists("mreadDNAStringSet")) {mreadDNAStringSet=memoise::memoise(readDNAStringSet, cache = cachem::cache_mem(max_size = 10*1024 * 1024^2))}


# 1. Récupérer le genome
genome <- mreadDNAStringSet(paste0("~/projects/datashare/genomes/", species, "/", annotation, "/", version, "/Sequence/WholeGenomeFasta/genome.fa"))
names(genome) <- sub(" .*", "", names(genome))  # clean des noms (important)

# 2. Récupérer tous les PWM humains depuis JASPAR
opts <- list(species=9606, all_versions=FALSE)  # humain
jaspar_matrices <- TFBSTools::getMatrixSet(JASPAR2022::JASPAR2022, opts)

# idx_peak="mrg1.00e-99_peak_3" 
# foo = call_jaspar(idx_peak=idx_peak, genome=genome, jaspar_matrices=jaspar_matrices, peaks=peaks)
# call_jaspar("mrg1.00e-99_peak_7690", genome=genome, jaspar_matrices=jaspar_matrices, peaks=peaks)
call_jaspar = function(idx_peak, genome, jaspar_matrices, peaks) {
  # idx_peak <- peak
  # idx_peak = rownames(peaks)[1]
  print(idx_peak)
  # 3. la sequence
  chr   <- peaks[idx_peak,1]
  start <- peaks[idx_peak,2]
  end   <- peaks[idx_peak,3]
  seq <- as.character(subseq(genome[[chr]], start=start, end=end))

  # 4. Scanner la séquence pour chaque motif
  all_tfbs <- lapply(names(jaspar_matrices), function(tf_name) {
    # tf_name = "MA1723.1" # names(jaspar_matrices)[1]
    # tf_name = "MA0139.1"
    # tf_name = "MA1929.1"
    # tf_name = "MA1930.1"
    pwm <- jaspar_matrices[[tf_name]]
    tfbs <- TFBSTools::searchSeq(toPWM(pwm), seq, min.score=0, strand="*")  # seuil 85% pour réduire faux positifs
    if(length(tfbs) > 0) {
      # print(tf_name)
      ret = data.frame(
        TF = name(pwm),
        start = start(tfbs),
        end = end(tfbs),
        strand = strand(tfbs),
        # pv = pvalues(tfbs, type="sampling"),
        score = score(tfbs)
      )
      ret = ret[order(-ret$score),]
      ret = ret[1,]
      return(ret)
    } else {
      NULL
    }
  })
  tfbs_df <- do.call(rbind, all_tfbs)
  return(tfbs_df)
}
# stop("EFN")
```


```{r peaks_jaspar, eval=JASPARPEAKS}
cl_bs = parallel::makeCluster(nb_cores/2, type="FORK")
tfbs = parallel::parApply(cl_bs, t(t(rownames(peaks))), 1, call_jaspar, genome=genome, jaspar_matrices=jaspar_matrices, peaks=peaks)
parallel::stopCluster(cl_bs)
# foo = call_jaspar(idx_peak="mrg1.00e-99_peak_3", genome=genome, jaspar_matrices=jaspar_matrices, peaks=peaks)
# tfbs = epimedtools::monitored_apply(t(t(rownames(peaks)[1:10])), 1, call_jaspar, genome=genome, jaspar_matrices=jaspar_matrices, peaks=peaks)
names(tfbs) = rownames(peaks)
saveRDS(tfbs, paste0(prefix2, "_tfbs.rds"))
```

```{r tfbs_stats}
tfbs = readRDS(paste0(prefix2, "_tfbs.rds"))
tf_scores = sapply(tfbs, "[[", "score")
rownames(tf_scores) = tfbs[[1]]$TF

layout(matrix(1:4,2), respect=TRUE)
den = density(tf_scores)
plot(den)
# apply(tf_scores, 1, function(tf_score) {
# tf_scores = sapply(tfbs, "[[", "score")
#   lines(density(tf_score, bw=den$bw))
# })

score_means = apply(tf_scores, 1, mean)
den = density(score_means)
plot(den)

len = tfbs[[1]][,3] - tfbs[[1]][,2]
den = density(len)
plot(den)

custom_hm(len, score_means, nbin=30, xlab="TFBS length", ylab="TFBS av. scores")
```

```{r tfbs_clusters}
pca = prcomp(tf_scores, scale=TRUE)
v = pca$sdev * pca$sdev
p = v / sum(v) * 100
# --- MCLUST sur les scores PCA ---
library(mclust)
mc = mclust::Mclust(pca$x[,2:6], G=5)
cl = mc$classification
layout(matrix(1:6,2), respect=TRUE)
# layout(matrix(1:2,1), respect=TRUE)
barplot(p, ylab="% of variance explained", xlab="components")

for (i in 1:5) {
  j = i+1
  plot(
    pca$x[,i], pca$x[,j],
    xlab=paste0("PC", i, "(", signif(p[i], 3), "%)"),
    ylab=paste0("PC", j, "(", signif(p[j], 3), "%)"),
    col=cl,          # <- coloration par cluster
    pch=16
  )
}

trecre = sort(names(cl)[cl==5])
foo = pca$x[trecre,4]
tre = names(foo)[foo>0] # TPA Response Element
cre = names(foo)[foo<0] # cAMP Response Element

col = trecre
names(col) = col
col[tre] = 1
col[cre] = 2

data = tf_scores[rownames(tf_scores)[rownames(tf_scores)%in%trecre],]
pca = prcomp(data, scale=TRUE)
v = pca$sdev * pca$sdev
p = v / sum(v) * 100
# --- MCLUST sur les scores PCA ---
library(mclust)
mc = mclust::Mclust(pca$x[,2:6], G=5)
cl = mc$classification
cl = as.numeric(col[rownames(pca$x)])
layout(matrix(1:6,2), respect=TRUE)
# layout(matrix(1:2,1), respect=TRUE)
barplot(p, ylab="% of variance explained", xlab="components")

for (i in 1:5) {
  j = i+1
  plot(
    pca$x[,i], pca$x[,j],
    xlab=paste0("PC", i, "(", signif(p[i], 3), "%)"),
    ylab=paste0("PC", j, "(", signif(p[j], 3), "%)"),
    col=cl,          # <- coloration par cluster
    pch=16
  )
}



data = tf_scores[rownames(tf_scores)[!rownames(tf_scores)%in%trecre],]
pca = prcomp(data, scale=TRUE)
v = pca$sdev * pca$sdev
p = v / sum(v) * 100
# --- MCLUST sur les scores PCA ---
library(mclust)
mc = mclust::Mclust(pca$x[,2:6], G=5)
cl = mc$classification
layout(matrix(1:6,2), respect=TRUE)
# layout(matrix(1:2,1), respect=TRUE)
barplot(p, ylab="% of variance explained", xlab="components")

for (i in 1:5) {
  j = i+1
  plot(
    pca$x[,i], pca$x[,j],
    xlab=paste0("PC", i, "(", signif(p[i], 3), "%)"),
    ylab=paste0("PC", j, "(", signif(p[j], 3), "%)"),
    col=cl,          # <- coloration par cluster
    pch=16
  )
}

homeobox = sort(names(cl)[cl==5]) # Homeobox

data = tf_scores[rownames(tf_scores)[!rownames(tf_scores)%in%c(trecre, homeobox)],]
pca = prcomp(data, scale=TRUE)
v = pca$sdev * pca$sdev
p = v / sum(v) * 100
# --- MCLUST sur les scores PCA ---
library(mclust)
mc = mclust::Mclust(pca$x[,2:6], G=5)
cl = mc$classification
layout(matrix(1:6,2), respect=TRUE)
# layout(matrix(1:2,1), respect=TRUE)
barplot(p, ylab="% of variance explained", xlab="components")

for (i in 1:5) {
  j = i+1
  plot(
    pca$x[,i], pca$x[,j],
    xlab=paste0("PC", i, "(", signif(p[i], 3), "%)"),
    ylab=paste0("PC", j, "(", signif(p[j], 3), "%)"),
    col=cl,          # <- coloration par cluster
    pch=16
  )
}

ets = sort(names(cl)[cl==5]) # Homeobox


data = tf_scores[rownames(tf_scores)[!rownames(tf_scores)%in%c(trecre, homeobox, ets)],]
pca = prcomp(data, scale=TRUE)
v = pca$sdev * pca$sdev
p = v / sum(v) * 100
# --- MCLUST sur les scores PCA ---
library(mclust)
mc = mclust::Mclust(pca$x[,2:6], G=5)
cl = mc$classification
layout(matrix(1:6,2), respect=TRUE)
# layout(matrix(1:2,1), respect=TRUE)
barplot(p, ylab="% of variance explained", xlab="components")

for (i in 1:5) {
  j = i+1
  plot(
    pca$x[,i], pca$x[,j],
    xlab=paste0("PC", i, "(", signif(p[i], 3), "%)"),
    ylab=paste0("PC", j, "(", signif(p[j], 3), "%)"),
    col=cl,          # <- coloration par cluster
    pch=16
  )
}

tfbsoi = sort(names(cl)[cl==1]) # Homeobox
metabo = tfbsoi[pca$x[tfbsoi,3]>0 & pca$x[tfbsoi,4]<0]

data = tf_scores[rownames(tf_scores)[!rownames(tf_scores)%in%c(trecre, homeobox, ets, metabo)],]
pca = prcomp(data, scale=TRUE)
v = pca$sdev * pca$sdev
p = v / sum(v) * 100
# --- MCLUST sur les scores PCA ---
library(mclust)
mc = mclust::Mclust(pca$x[,2:6], G=5)
cl = mc$classification
layout(matrix(1:6,2), respect=TRUE)
# layout(matrix(1:2,1), respect=TRUE)
barplot(p, ylab="% of variance explained", xlab="components")

for (i in 1:5) {
  j = i+1
  plot(
    pca$x[,i], pca$x[,j],
    xlab=paste0("PC", i, "(", signif(p[i], 3), "%)"),
    ylab=paste0("PC", j, "(", signif(p[j], 3), "%)"),
    col=cl,          # <- coloration par cluster
    pch=16
  )
}

```

Filtering `tf_scores`


```{r}
deseq2_genes = read.table("../GSE193141/04_deseq2/tables/MvsF.complete.txt", header=TRUE)
rownames(deseq2_genes) = deseq2_genes[,1]
deseq2_genes = deseq2_genes[,grep("norm", colnames(deseq2_genes))]
colnames(deseq2_genes) = substr(colnames(deseq2_genes), 6, 200)
deseq2_genes = as.matrix(deseq2_genes)
# remove not expressd genes
deseq2_genes = deseq2_genes[!apply(deseq2_genes==0, 1, all),]

foo = rownames(tf_scores)[!rownames(tf_scores) %in% deseq2_genes]
split_dimer = unique(unlist(strsplit(foo, "::")))
split_dimer[!split_dimer %in% deseq2_genes]
 # [1] "EWSR1-FLI1" "TBXT"      
plot(density(deseq2_genes["T",]))

notnull_genes = rownames(deseq2_genes)

simple_tfs = rownames(tf_scores)[rownames(tf_scores) %in% notnull_genes]
max_tf = apply(deseq2_genes[simple_tfs, ], 1, max)
plot(sort(max_tf), ylim=c(0,100))
simple_tfs = names(max_tf)[max_tf>=20]

dimer_tfs = foo[grep("::", foo)]
split_dimer = unique(unlist(strsplit(dimer_tfs, "::")))
split_dimer = split_dimer[split_dimer %in% notnull_genes]
max_tf = apply(deseq2_genes[split_dimer, ], 1, max)
plot(sort(max_tf), ylim=c(0,1000))
blacklist_tfs = names(max_tf)[max_tf<20]
dimer_tfs = dimer_tfs[-grep(paste0(blacklist_tfs, collapse="|"), foo)]

tf_scores_filtered = tf_scores[rownames(tf_scores)%in%c(simple_tfs, dimer_tfs),]
dim(tf_scores_filtered)
tf_scores = tf_scores_filtered
```



## Peak counts

```{r peaks_featureCounts, eval=COUNTPEAKS}
# Counts
feat_saf = peaks[, c(4,1:3, 6)]
# feat_saf = peaks[1:100,c(4,1:3, 6)]
colnames(feat_saf) = c("GeneID", "Chr", "Start", "End", "Strand")
# feat_saf$Start = feat$tss
# head(feat_saf)
# feat_saf$Start = feat$tss - , 800, 1500) 
# if (sum(feat_saf$Start<1)!=0) {
#   feat_saf[feat_saf$Start<1,]$Start = 1
# } 
# feat_saf$End = feat$tss + ifelse(feat_saf$Strand=="-", 800, 1500)  
# head(feat_saf)
saf_filename = paste0(prefix2, ".saf")
write.table(feat_saf, file=saf_filename, sep="\t", quote=FALSE,row.names=FALSE, col.names=FALSE)
cmd = "/summer/epistorage/miniconda3/envs/subread_env/bin/featureCounts"
cmd = "featureCounts"
args = paste0(
  "-O ", # allow multiple overlap
  "-p -a ", saf_filename, " -F SAF -s 0 -Q ", mmq, " -T 16 -o ", prefix2, "_count_", mmq, ".txt ", 
  paste0("/home/chuffarf/projects/datashare/", gse, "/", e$gsm_atac, "_end-to-end_trimno_bowtie2_", species, "_", annotation, "_", version, "_srt.bam ", collapse=" "),
  ""
)
print(paste(cmd, args))
system2(cmd, args)
system2("cat", paste0(prefix2, "_count_", mmq, ".txt.summary"))
```

```{r raw_counts_peaks}
raw_counts = read.table(paste0(prefix2, "_count_", mmq, ".txt"), header=TRUE)
rownames(raw_counts) = raw_counts[,1]
raw_counts = raw_counts[,-(1:6)]
colnames(raw_counts) = sub(".*?(GSM[^_]*)_.*", "\\1", colnames(raw_counts))
head(raw_counts[,1:10])
tail(raw_counts[,1:10])
dim( raw_counts)
```

```{r norm_counts_peaks, eval=TRUE}
if (!all(e$gsm_atac %in% colnames(raw_counts))) {stop("problem e$gsm_atac vs. colnames(raw_counts)")}
countData = raw_counts[,e$gsm_atac]
colnames(countData) = rownames(e)
colData = e[colnames(countData), c("age", "gender")]
# Vérifie que les noms de colonnes correspondent
if (!all(colnames(countData) == rownames(colData))) {stop("probleme data vs. exp_grp")}

dds <- DESeq2::DESeqDataSetFromMatrix(
  countData = countData,
  colData = colData,
  design = ~ gender + age
)

# dds <- DESeq2::DESeq(dds)
# res_age <- DESeq2::results(dds, name = "age")

dds = DESeq2::estimateSizeFactors(dds)

norm_counts_peaks = DESeq2::counts(dds, normalized=TRUE)
head(norm_counts_peaks[,1:10])
dim(norm_counts_peaks)
```

```{r pca_norm_counts_peaks}
data = as.matrix(t(norm_counts_peaks))
pca = prcomp(data, scale=TRUE)
v = pca$sdev * pca$sdev
p = v / sum(v) * 100
# # --- MCLUST sur les scores PCA ---
# library(mclust)
# mc = mclust::Mclust(pca$x[,2:6], G=5)
# cl = mc$classification
# col = cl

col = as.numeric(e[rownames(pca$x),]$gender)
layout(matrix(1:6,2), respect=TRUE)
# layout(matrix(1:2,1), respect=TRUE)
barplot(p, ylab="% of variance explained", xlab="components")

for (i in 1:5) {
  j = i+1
  plot(
    pca$x[,i], pca$x[,j],
    xlab=paste0("PC", i, "(", signif(p[i], 3), "%)"),
    ylab=paste0("PC", j, "(", signif(p[j], 3), "%)"),
    col=col,
    pch=16
  )
}

for (key in c("age", "gender")) {
 layout(matrix(1:6,2), respect=TRUE)
  plot(sapply(1:ncol(pca$x), function(i) {
    x =  e[rownames(pca$x),key]
    y = pca$x[,i]
    m = lm(y~x)
    summary(m)$r.squared*100
  }), main=key, xlab="PCs", ylab="R^2 (%)")
  for (i in 1:5) {
    x =  e[rownames(pca$x),key]
    y = pca$x[,i]
    m = lm(y~x)
    xlab=key
    ylab=paste0("PC", i, "(", signif(p[i], 3), "%)")
    main=paste0("R^2=", signif(summary(m)$r.squared, 3)*100, "%")
    if (is.numeric(x)) {
      plot(x, y, xlab=xlab, ylab=ylab, main=main, pch=16)      
      abline(m, col=2)
    } else {
      boxplot(y~x, xlab=xlab, ylab=ylab, main=main)      
    }
  }
}
```




# RNA

```{r norm_counts_rna}
# awk '$3 == "transcript"' /home/chuffarf/projects/datashare/genomes/Homo_sapiens/T2T/CHM13v2.0/Annotation/Genes/genes.gtf |  awk 'OFS="\t" {if ($3=="transcript") {print $1,$4-1,$5,$10,$6,$7}}' | tr -d '";' | awk '!seen[$4]++'  > /home/chuffarf/projects/datashare/genomes/Homo_sapiens/T2T/CHM13v2.0/Annotation/Genes/genes.bed
# awk 'BEGIN{FS=OFS="\t"}
# {
#     if (match($9,/gene_name "[^"]+"/)) {
#         gene=substr($9,RSTART+11,RLENGTH-12)
#     } else if (match($9,/gene_id "[^"]+"/)) {
#         gene=substr($9,RSTART+9,RLENGTH-10)
#     } else next

#     if (!(gene in min) || $4 < min[gene]) min[gene]=$4
#     if (!(gene in max) || $5 > max[gene]) max[gene]=$5
#     chr[gene]=$1
#     strand[gene]=$7
# }
# END{
#     for (g in min)
#         print chr[g], min[g]-1, max[g], g, ".", strand[g]
# }' /home/chuffarf/projects/datashare/genomes/Homo_sapiens/UCSC/hg38/Annotation/Genes/genes.gtf | sort -k1,1 -k2,2n > /home/chuffarf/projects/datashare/genomes/Homo_sapiens/UCSC/hg38/Annotation/Genes/genes.bed

genes_bed = paste0("~/projects/datashare/genomes/", species, "/", annotation, "/", version, "/Annotation/Genes/genes.bed")
genes_bed
genes = read.table(genes_bed)
genes = genes[nchar(genes[,1])<=5,]
rownames(genes) = genes[,4]
head(genes)
dim(genes)

norm_counts_rna = read.table("../GSE193141/04_deseq2/tables/MvsF.complete.txt", header=TRUE)
rownames(norm_counts_rna) = norm_counts_rna[,1]
norm_counts_rna = norm_counts_rna[,grep("norm", colnames(norm_counts_rna))]
colnames(norm_counts_rna) = substr(colnames(norm_counts_rna), 6, 200)
norm_counts_rna = as.matrix(norm_counts_rna)

if (!all(e$gsm_rna %in% colnames(norm_counts_rna))) {stop("problem e$gsm_rna vs. colnames(norm_counts_rna)")}
norm_counts_rna = norm_counts_rna[,e$gsm_rna]
colnames(norm_counts_rna) = rownames(e)

# remove not expressd genes
norm_counts_rna = norm_counts_rna[!apply(norm_counts_rna==0, 1, all),]
# sum( rownames(norm_counts_rna) %in% rownames(genes))
# sum(!rownames(norm_counts_rna) %in% rownames(genes))
genes = genes[rownames(norm_counts_rna)[rownames(norm_counts_rna) %in% rownames(genes)],]
norm_counts_rna = norm_counts_rna[rownames(genes),]

head(norm_counts_rna[,1:10])
dim(norm_counts_rna)
```


```{r pca_norm_counts_rna}
data = as.matrix(t(norm_counts_rna))
pca = prcomp(data, scale=TRUE)
v = pca$sdev * pca$sdev
p = v / sum(v) * 100
# # --- MCLUST sur les scores PCA ---
# library(mclust)
# mc = mclust::Mclust(pca$x[,2:6], G=5)
# cl = mc$classification
# col = cl

col = as.numeric(e[rownames(pca$x),]$gender)
layout(matrix(1:6,2), respect=TRUE)
# layout(matrix(1:2,1), respect=TRUE)
barplot(p, ylab="% of variance explained", xlab="components")

for (i in 1:5) {
  j = i+1
  plot(
    pca$x[,i], pca$x[,j],
    xlab=paste0("PC", i, "(", signif(p[i], 3), "%)"),
    ylab=paste0("PC", j, "(", signif(p[j], 3), "%)"),
    col=col,
    pch=16
  )
}

for (key in c("age", "gender")) {
 layout(matrix(1:6,2), respect=TRUE)
  plot(sapply(1:ncol(pca$x), function(i) {
    x =  e[rownames(pca$x),key]
    y = pca$x[,i]
    m = lm(y~x)
    summary(m)$r.squared*100
  }), main=key, xlab="PCs", ylab="R^2 (%)")
  for (i in 1:5) {
    x =  e[rownames(pca$x),key]
    y = pca$x[,i]
    m = lm(y~x)
    xlab=key
    ylab=paste0("PC", i, "(", signif(p[i], 3), "%)")
    main=paste0("R^2=", signif(summary(m)$r.squared, 3)*100, "%")
    if (is.numeric(x)) {
      plot(x, y, xlab=xlab, ylab=ylab, main=main, pch=16)      
      abline(m, col=2)
    } else {
      boxplot(y~x, xlab=xlab, ylab=ylab, main=main)      
    }
  }
}





# mat = t(foo)
# mat = mat[,apply(!is.na(mat), 2, all)]
data = log2(as.matrix(t(norm_counts_rna))+1)
data = as.matrix(t(norm_counts_rna))+1
pca = prcomp(data, scale=TRUE)
v = pca$sdev * pca$sdev
p = v / sum(v) * 100
palette("default")

# Distance de Mahalanobis
d2 <- mahalanobis(pca$x[,1:2], center=colMeans(pca$x[,1:2]), cov=cov(pca$x[,1:2]))
# Seuil correspondant à 95% (5% à l'extérieur)
threshold <- qchisq(0.99, df = 2)
# Points extérieurs
outliers <- d2 > threshold
outliers = names(outliers)[outliers]

# # gsms = gsms[!gsms%in%outliers]
# e = e[!e$gsm_atac%in%outliers,]

layout(matrix(1:2,1), respect=TRUE)
# layout(matrix(1:2,1), respect=TRUE)
barplot(p, ylab="% of variance explained", xlab="components")
i = 1
j = i+1
plot(pca$x[,i], pca$x[,j], 
  xlab=paste0("PC", i, "(", signif(p[i], 3), "%)"), 
  ylab=paste0("PC", j, "(", signif(p[j], 3), "%)"), 
  col=(rownames(pca$x)%in%outliers)+1,
  pch=16
)  


den = density(log2(norm_counts_rna+1))
plot(den, ylim=c(0,.6), xlim=c(-1, 5))
for (i in rownames(data)) {
  den2 = density(log2(norm_counts_rna[,i]+1), bw=den$bw)
  lines(den2, col=adjustcolor(i%in%outliers+1, alpha.f=.3))
}
for (i in outliers) {
  den2 = density(log2(norm_counts_rna[,i]+1), bw=den$bw)
  lines(den2, col=2)
}

raw_counts_rna = read.table("../GSE193141/04_deseq2/tables/MvsF.complete.txt", header=TRUE)
rownames(raw_counts_rna) = raw_counts_rna[,1]
cn = colnames(raw_counts_rna)[grep("norm", colnames(raw_counts_rna))]
cn = substr(cn, 6, 100)
raw_counts_rna = raw_counts_rna[,cn]
if (!all(e$gsm_rna %in% colnames(raw_counts_rna))) {stop("problem e$gsm_rna vs. colnames(raw_counts_rna)")}
raw_counts_rna = raw_counts_rna[,e$gsm_rna]
colnames(raw_counts_rna) = rownames(e)
raw_counts_rna = as.matrix(raw_counts_rna)

den = density(log2(raw_counts_rna+1))
plot(den, ylim=c(0,1), xlim=c(-1, 15))
# plot(den)
for (i in rownames(data)) {
  den2 = density(log2(raw_counts_rna[,i]+1), bw=den$bw)
  lines(den2, col=adjustcolor(i%in%outliers+1, alpha.f=.3))
}
for (i in outliers) {
  den2 = density(log2(raw_counts_rna[,i]+1), bw=den$bw)
  lines(den2, col=2)
}

den = density(log2(raw_counts_rna+1))
plot(den, ylim=c(0,1), xlim=c(-1, 2))
# plot(den)
for (i in rownames(data)) {
  den2 = density(log2(raw_counts_rna[,i]+1), bw=den$bw)
  lines(den2, col=adjustcolor(i%in%outliers+1, alpha.f=.3))
}
for (i in outliers) {
  den2 = density(log2(raw_counts_rna[,i]+1), bw=den$bw)
  lines(den2, col=2)
}

# update exp_grp and norm_counts_peaks
e = e[!rownames(e)%in%outliers,]
norm_counts_peaks = norm_counts_peaks[,rownames(e)]
norm_counts_rna = norm_counts_rna[,rownames(e)]

# remove not expressd genes
norm_counts_rna = norm_counts_rna[!apply(norm_counts_rna==0, 1, all),]
# sum( rownames(norm_counts_rna) %in% rownames(genes))
# sum(!rownames(norm_counts_rna) %in% rownames(genes))
genes = genes[rownames(norm_counts_rna)[rownames(norm_counts_rna) %in% rownames(genes)],]
norm_counts_rna = norm_counts_rna[rownames(genes),]

head(norm_counts_rna[,1:10])
dim(norm_counts_rna)


```


# Joint


```{r genes_peaks_joint}
# stop("EFN")
head(genes)
dim(genes)

library(GenomicRanges)

# Peaks
gr_peaks <- GRanges(
  seqnames = peaks$chrom,
  ranges = IRanges(start = peaks$chromStart + 1, end = peaks$chromEnd),
  peak = peaks$name
)

# Genes → TSS
# ajouter noms de colonnes
colnames(genes) <- c("chr", "start", "end", "gene", "score", "strand")

# définir le TSS selon le strand
tss_pos <- ifelse(genes$strand == "+", genes$start, genes$end)

gr_tss <- GRanges(
  seqnames = genes$chr,
  ranges = IRanges(start = tss_pos, end = tss_pos),
  strand = genes$strand,
  gene = genes$gene
)

max_dist = 500e6 # full chromosomes
# max_dist <- .1e6

# fenêtre autour des TSS
gr_tss_window <- resize(gr_tss, width = 2 * max_dist + 1, fix = "center")

hits <- findOverlaps(gr_peaks, gr_tss_window)

peak_hits <- gr_peaks[queryHits(hits)]
tss_hits  <- gr_tss[subjectHits(hits)]

distances <- distance(peak_hits, tss_hits)

joint <- data.frame(
  peak = mcols(peak_hits)$peak,
  gene    = mcols(tss_hits)$gene,
  chr     = as.character(seqnames(peak_hits)),
  peak_start = start(peak_hits),
  peak_end   = end(peak_hits),
  tss        = start(tss_hits),
  strand     = as.character(strand(tss_hits)),
  distance   = distances
)

head(joint)
dim(joint)
# stop("EFN")
```

# Mediation


```{r med_func}
med_func = function(j, norm_counts_peaks, norm_counts_rna, e, LF=NULL) {
  if (!all(colnames(norm_counts_rna) == colnames(norm_counts_peaks))) {stop("probleme colnames(norm_counts_rna) vs. colnames(norm_counts_peaks)")}
  if (!all(rownames(e) == colnames(norm_counts_peaks))) {stop("probleme rownames(e) vs. colnames(norm_counts_peaks)")}

  # print(j)
  # j = joint[1,]
  idx_peak = j[[1]]
  idx_gene = j[[2]]
  # print(paste0(idx_peak, " ", idx_gene))
  M = log2(norm_counts_peaks[idx_peak,]+1)
  X = e$age
  Y = log2(norm_counts_rna[idx_gene,]+1)
  C = e$gender
  at = mean(M)
  rn = mean(Y)
  m  = lm(Y ~ X + C)
  m1 = lm(M ~ X + C)
  m2 = lm(Y ~ M + X + C)
  if (!is.null(LF)) {
    if (!all(rownames(LF) == colnames(norm_counts_peaks))) {stop("probleme rownames(LF) vs. colnames(norm_counts_peaks)")}
    # LF = pca$x[,1:2]
    m  = lm(Y ~ X + C + LF)
    m1 = lm(M ~ X + C + LF)
    m2 = lm(Y ~ M + X + C + LF)
  }
  a = m1$coef[[2]]
  b = m2$coef[[2]]
  c = m$coef[[2]]
  cp = m2$coef[[3]]
  pv  = summary(m )$coefficients[2,4]
  pv1 = summary(m1)$coefficients[2,4]
  pv2 = summary(m2)$coefficients[2,4]
  pv3 = summary(m2)$coefficients[3,4]
  pmax2 = max(pv1, pv2)^2
  ret = c(
    # i     = i     ,
    # j     = j     , 
    RNA   = rn    ,
    ATAC  = at    ,
    ADE   = cp    ,
    ACME  = a*b   ,
    TE    = c     ,
    a     = a     ,   
    b     = b     ,   
    c     = c     ,  
    cp    = cp    ,   
    pv    = pv    ,
    pv1   = pv1   ,                    
    pv2   = pv2   ,                    
    pv3   = pv3   ,                    
    pmax2 = pmax2       
  )
  return(ret)
}
```

```{r mediation, eval=MEDRES}
### LF with PCA
# data = log2(as.matrix(t(norm_counts_rna))+1)
data = log2(as.matrix(t(norm_counts_peaks))+1)
pca_peaks = pca = prcomp(data, scale=TRUE)
LF = pca_peaks$x[,1:2]

### LF with lfmm
data = log2(as.matrix(t(norm_counts_peaks))+1)
X = c(age=e[rownames(data), "age"])
Y = data
lfmm_fit = lfmm::lfmm_ridge(Y=Y, X=X, K=3,  lambda=1e-5)
U = lfmm_fit$U # latent factors
V = lfmm_fit$V # loadings
B = lfmm_fit$B # 
rownames(U) = rownames(data)
rownames(V) = colnames(data)
LF = U[,c(1,1)]

# LF = NULL

# medres = epimedtools::monitored_apply(joint[,1:2], 1, med_func, norm_counts_peaks=norm_counts_peaks, norm_counts_rna=norm_counts_rna, e=e, LF=LF)
cl_bs = parallel::makeCluster(nb_cores, type="FORK")
medres = parallel::parApply(cl_bs, joint[,1:2], 1, med_func, norm_counts_peaks=norm_counts_peaks, norm_counts_rna=norm_counts_rna, e=e, LF=LF)
parallel::stopCluster(cl_bs)

medres = t(medres)
medres = cbind(joint, medres)
medres$pmax2adj = p.adjust(medres$pmax2, method="BH")
saveRDS(medres, paste0(prefix2, "_medres.rds"))
# medres = do.call(rbind, medres)
```




```{r medres}
# stop("EFN")
res = readRDS(paste0(prefix2, "_medres.rds"))
res = res[res$peak %in% rownames(peaks) & res$gene %in% rownames(genes),]
res = res[res$RNA != 0,]

joint$key = paste0(joint[,"peak"], "-", joint[,"gene"])
sum(duplicated(joint$key))

res$key = paste0(res[,"peak"], "-", res[,"gene"])
sum(duplicated(res$key))

all(joint$key %in% res$key)
all(res$key %in% joint$key)

# j = joint[!joint$key %in% res$key,][1,1:2]

# layout(1, respect=TRUE)
# idx = !is.na(res$pmax2) & res$pmax2adj < 0.05
# # plot(-log10(res$pmax2adj), col=(rownames(res) %in% idx)+1)
# plot(res$ACME, -log10(res$pmax2adj), col=idx+1, pch=".")
# idx2 = !is.na(res$pmax2) & -log10(res$pmax2adj)>10
# text(res[idx2,]$ACME, -log10(res[idx2,]$pmax2adj), res[idx2,]$gene)

# custom_hm(res$ACME, -log10(res$pmax2adj), LOGCOUNT=TRUE)

```

```{r, echo=TRUE, results="verbatim"} 
dim(joint)
dim(res)
# stop("EFN")
```


# Local genes / peaks interactions

```{r ebf1_bis}
stop("EFN")
gene = "EBF1"
peak = "mrg1.00e-99_peak_61474"
peaks[peak,]
subres = plot_hmmed(gene=gene, res)
fitMed = plot_mediationpkg(gene=gene, peak=peak)
enrich = plot_enrich(peaks=peak, tf_scores=tf_scores, res=res, main=peak)

gene = "EXD2"
peak = "mrg1.00e-99_peak_22472a"
peaks[peak,]
subres = plot_hmmed(gene=gene, res, delta=5e5)
fitMed = plot_mediationpkg(gene=gene, peak=peak)
enrich = plot_enrich(peaks=peak, tf_scores=tf_scores, res=res, main=peak)

gene = "OBSCN"
peak = "mrg1.00e-99_peak_7072"
peaks[peak,]
subres = plot_hmmed(gene=gene, res, delta=5e5)
fitMed = plot_mediationpkg(gene=gene, peak=peak)
enrich = plot_enrich(peaks=peak, tf_scores=tf_scores, res=res, main=peak)

# subres = plot_hmmed("TRABD2A", res)
# subres = plot_hmmed("GREM2", res)
# subres = plot_hmmed("CHP1", res)
# subres = plot_hmmed("GAL3ST4", res)
# subres = plot_hmmed("NOG", res)
# subres = plot_hmmed("DNAH6", res)
# subres = plot_hmmed("EBF1", res, delta=20e6)
# subres = plot_hmmed("JUN", res)
# subres = subres[!is.na(subres$pmax2)&subres$gene=="JUN", ]
# subres[order(subres$ACME), 1:6]
```


# UP/DOWN Grouping

```{r eval=TRUE}
res = res[res$distance<.5e6,]
res = res[res$distance<5000,]
length(unique(res$gene))
length(unique(res$peak))
dim(res)
subres = plot_hmmed(gene="EBF1", res)

idx = p.adjust(res$pv, method="BH") < 0.05
subres = res[idx,]
idx = subres$pmax2adj < 0.05
subres = subres[idx,]

genes_up = unique(subres[subres$c>=0,]$gene)
genes_dn = unique(subres[subres$c< 0,]$gene)
peaks_up = unique(subres[subres$a>=0,]$peak)
peaks_dn = unique(subres[subres$a< 0,]$peak)

layout(matrix(1:6,2), respect=TRUE)

plot(res[!duplicated(res$gene),]$c, -log10(res[!duplicated(res$gene),]$pv), pch=".")
points(subres[!duplicated(subres$gene),]$c, -log10(subres[!duplicated(subres$gene),]$pv), col=2)
# text(subres[!duplicated(subres$gene),]$c, -log10(subres[!duplicated(subres$gene),]$pv), subres[!duplicated(subres$gene),]$gene)

plot(res[!duplicated(res$peak),]$a, -log10(res[!duplicated(res$peak),]$pv1), pch=".")
points(subres[!duplicated(subres$peak),]$a, -log10(subres[!duplicated(subres$peak),]$pv1), col=2)
# text(subres[!duplicated(subres$gene),]$c, -log10(subres[!duplicated(subres$gene),]$pv), subres[!duplicated(subres$gene),]$gene)

plot(subres$ACME, -log10(subres$pmax2adj), pch=".", main="genes_up"); points(subres[subres$gene%in%genes_up,]$ACME, -log10(subres[subres$gene%in%genes_up,]$pmax2adj), col=2)
plot(subres$ACME, -log10(subres$pmax2adj), pch=".", main="genes_dn"); points(subres[subres$gene%in%genes_dn,]$ACME, -log10(subres[subres$gene%in%genes_dn,]$pmax2adj), col=3)
plot(subres$ACME, -log10(subres$pmax2adj), pch=".", main="peaks_up"); points(subres[subres$peak%in%peaks_up,]$ACME, -log10(subres[subres$peak%in%peaks_up,]$pmax2adj), col=4)
plot(subres$ACME, -log10(subres$pmax2adj), pch=".", main="peaks_dn"); points(subres[subres$peak%in%peaks_dn,]$ACME, -log10(subres[subres$peak%in%peaks_dn,]$pmax2adj), col=5)

layout(matrix(1:2,1), respect=TRUE)
enrich = plot_enrich(peaks=peaks_up, tf_scores=tf_scores, res=res, main="up")
enrich = plot_enrich(peaks=peaks_up, tf_scores=tf_scores, res=res, main="DOWN")

```


# Candidates

```{r}
idx = p.adjust(res$pv, method="BH") < 0.05
subres = res[idx,]
idx = subres$pmax2adj < 0.05
subres = subres[idx,]


rownames(subres) = subres$key
foo = subres[order(subres$pmax2adj), c("ACME", "pmax2adj"),]
idx = rownames(foo)[1:20]
layout(1, respect=TRUE)
plot(subres$ACME, -log10(subres$pmax2adj), pch=".", main="genes_oi"); 
points(subres[idx,]$ACME, -log10(subres[idx,]$pmax2adj), col=2)
text(subres[idx,]$ACME, -log10(subres[idx,]$pmax2adj), subres[idx,]$gene)
for (i in idx) {
  gene = subres[i,]$gene
  peak = subres[i,]$peak
  hmmed = plot_hmmed(gene=gene, res, delta=.5e6)
  fitMed = plot_mediationpkg(gene=gene, peak=peak)
  enrich = plot_enrich(peaks=peak, tf_scores=tf_scores, res=res, main=peak)
}

```




















# ATAC~age

```{r enrich_atac}

idx = !duplicated(res$peak)
subres = res[idx,]
plot(subres$a, -log10(subres$pv1), xlab="a", ylab="pva")

bg_peaks = unique(res$peak)
tf_bg = apply(tf_scores[, bg_peaks], 1, median)

idx_peaks_oi = subres[subres$pv1 < 0.05&subres$a < 0,]$peak
enrich = plot_enrich(peaks=idx_peaks_oi, tf_scores=tf_scores, res=res, main="DOWN")

idx_peaks_oi = subres[subres$pv1 < 0.05&subres$a > 0,]$peak
enrich = plot_enrich(peaks=idx_peaks_oi, tf_scores=tf_scores, res=res, main="UP")

```

```{r pca4LF, eval=FALSE}
data = log2(as.matrix(t(norm_counts_rna))+1)
pca_rna = prcomp(data, scale=TRUE)
data = log2(as.matrix(t(norm_counts_peaks))+1)
pca_peaks = pca = prcomp(data, scale=TRUE)
v = pca$sdev * pca$sdev
p = v / sum(v) * 100
# # --- MCLUST sur les scores PCA ---
# library(mclust)
# mc = mclust::Mclust(pca$x[,2:6], G=5)
# cl = mc$classification
# col = cl

col = as.numeric(e[rownames(pca$x),]$gender)
layout(matrix(1:6,2), respect=TRUE)
# layout(matrix(1:2,1), respect=TRUE)
barplot(p, ylab="% of variance explained", xlab="components")

for (i in 1:5) {
  j = i+1
  plot(
    pca$x[,i], pca$x[,j],
    xlab=paste0("PC", i, "(", signif(p[i], 3), "%)"),
    ylab=paste0("PC", j, "(", signif(p[j], 3), "%)"),
    col=col,
    pch=16
  )
}

for (key in c("gender", "age")) {
 layout(matrix(1:6,2), respect=TRUE)
  plot(sapply(1:ncol(pca$x), function(i) {
    x =  e[rownames(pca$x),key]
    y = pca$x[,i]
    m = lm(y~x)
    summary(m)$r.squared*100
  }), main=key, xlab="PCs", ylab="R^2 (%)")
  for (i in 1:5) {
    x =  e[rownames(pca$x),key]
    y = pca$x[,i]
    m = lm(y~x)
    xlab=key
    ylab=paste0("PC", i, "(", signif(p[i], 3), "%)")
    main=paste0("R^2=", signif(summary(m)$r.squared, 3)*100, "%")
    if (is.numeric(x)) {
      plot(x, y, xlab=xlab, ylab=ylab, main=main, pch=16)      
      abline(m, col=2)
    } else {
      boxplot(y~x, xlab=xlab, ylab=ylab, main=main)      
    }
  }
}


res = readRDS(paste0(prefix2, "_medres.rds"))
dim(res)
res = res[res$peak %in% rownames(peaks) & res$gene %in% rownames(genes),]
dim(res)
res = res[res$RNA != 0,]
dim(res)

idx = res$peak %in% res[res$gene=="EBF1",]$peak
subres = res[idx,]


dev.new(width=9, height=9)
subres = plot_hmmed("EBF1", res, delta=20e6)

for (i in 1:3) {
  LF = cbind(pca_rna$x[,1], pca_peaks$x[,c(1:2)])
  LF = pca_peaks$x[,c(1:2)]
  # medreslf = epimedtools::monitored_apply(subres[,1:2], 1, med_func, norm_counts_peaks=norm_counts_peaks, norm_counts_rna=norm_counts_rna, e=e, LF=LF)
  cl_bs = parallel::makeCluster(10, type="FORK")
  medres2 =      parallel::parApply(cl_bs, subres[,1:2], 1, med_func, norm_counts_peaks=norm_counts_peaks, norm_counts_rna=norm_counts_rna, e=e, LF=LF)
  parallel::stopCluster(cl_bs)

  medres2 = t(medres2)
  medres2 = cbind(subres[,1:9], medres2)
  medres2$pmax2adj = p.adjust(medres2$pmax2, method="BH")
  # medres = do.call(rbind, medres)

  # subres = plot_hmmed("EBF1", res)
  # dev.new()
  dev.new(width=9, height=9)
  plot_hmmed("SHANK1", medres2, delta=5e6)
}

plot_hmmed("SHANK1", res, delta=5e6)
```

```{r lfmm, eval=FALSE}
data = log2(as.matrix(t(norm_counts_peaks))+1)
# data = apply(data, 2, function(x) {
#   (x - mean(x)) / sd(x)
# })
# X = c(gender=as.numeric(e[rownames(data),"gender"]))
X = c(age=e[rownames(data), "age"])
# X = rep(1, nrow(data))
# X = cbind(gender=as.numeric(e[rownames(data),"gender"]),age=e[rownames(data),"age"])
Y = data
lfmm_fit = lfmm::lfmm_ridge(Y=Y, X=X, K=3,  lambda=1e-5)
# dev.new()
# lfmm_fit = lfmm::lfmm_ridge(Y=Y, X=X, K=2,  lambda=100000)
# lfmm_fit = lfmm::lfmm_lasso(Y=Y, X=X, K=17, lambda=0.5)
# Y = XB + UV + E

U = lfmm_fit$U # latent factors
V = lfmm_fit$V # loadings
B = lfmm_fit$B # 

a = res[!duplicated(res$peak),]$a
names(a) = res[!duplicated(res$peak),]$peak
a = a[rownames(B)]
plot(a, B, pch=".")
abline(a=0, b=1, col=2)



rownames(U) = rownames(data)
rownames(V) = colnames(data)

layout(matrix(1:18,3), respect=TRUE)
plot(apply(abs(V), 2, sum) )

plot(sapply(1:ncol(U), function(i) {
  key="age"
  x =  e[rownames(U),key]
  y = U[,i]
  m = lm(y~x)
  summary(m)$r.squared*100
}), main=key, xlab="Us", ylab="R^2 (%)")
for (i in 1:16) {
  # plot(density(abs(V[,i]), bw=.0003), xlim=c(0,0.015))
  # idx_peaks_oi = names(rev(sort(abs(V[,i]))))[1:4000]
  idx_peaks_oi = rownames(V)[abs(V[,i])>0.01]
  length(idx_peaks_oi)
  tf_signal = apply(tf_scores[, idx_peaks_oi], 1, median)
  enrich = sort(tf_signal - tf_bg, decreasing=TRUE)
  plot_enrich(enrich)

  # plot(density(U[,i]))
}

# for (key in c("gender", "age")) {
for (key in c("age")) {
 layout(matrix(1:18,3), respect=TRUE)
  plot(sapply(1:ncol(U), function(i) {
    x =  e[rownames(U),key]
    y = U[,i]
    m = lm(y~x)
    summary(m)$r.squared*100
  }), main=key, xlab="Us", ylab="R^2 (%)")
  for (i in 1:17) {
    x =  e[rownames(U),key]
    y = U[,i]
    m = lm(y~x)
    xlab=key
    ylab=paste0("U", i)
    main=paste0("R^2=", signif(summary(m)$r.squared, 3)*100, "%")
    if (is.numeric(x)) {
      plot(x, y, xlab=xlab, ylab=ylab, main=main, pch=16)      
      abline(m, col=2)
    } else {
      boxplot(y~x, xlab=xlab, ylab=ylab, main=main)      
    }
  }
}

```




























# UP/DOWN Grouping

```{r eval=FALSE}
res = res[res$distance<.5e6,]
res = res[res$distance<5000,]
length(unique(res$gene))
length(unique(res$peak))
dim(res)


idx = p.adjust(res$pv, method="BH") < 0.05
subres = res[idx,]
length(unique(subres$gene))
sum(subres$pmax2adj < 0.05)
idx = subres$pmax2adj < 0.05
subres = subres[idx,]
idx_peaks_oi = unique(subres$peak)
bg_peaks = unique(res$peak)
length(idx_peaks_oi)

genes_up = unique(subres[subres$c>=0,]$gene)
genes_dn = unique(subres[subres$c< 0,]$gene)
peaks_up = unique(subres[subres$a>=0,]$peak)
peaks_dn = unique(subres[subres$a< 0,]$peak)

layout(matrix(1:6,2), respect=TRUE)

plot(res[!duplicated(res$gene),]$c, -log10(res[!duplicated(res$gene),]$pv), pch=".")
points(subres[!duplicated(subres$gene),]$c, -log10(subres[!duplicated(subres$gene),]$pv), col=2)
# text(subres[!duplicated(subres$gene),]$c, -log10(subres[!duplicated(subres$gene),]$pv), subres[!duplicated(subres$gene),]$gene)

plot(res[!duplicated(res$peak),]$a, -log10(res[!duplicated(res$peak),]$pv1), pch=".")
points(subres[!duplicated(subres$peak),]$a, -log10(subres[!duplicated(subres$peak),]$pv1), col=2)
# text(subres[!duplicated(subres$gene),]$c, -log10(subres[!duplicated(subres$gene),]$pv), subres[!duplicated(subres$gene),]$gene)

plot(subres$ACME, -log10(subres$pmax2adj), pch=".", main="genes_up"); points(subres[subres$gene%in%genes_up,]$ACME, -log10(subres[subres$gene%in%genes_up,]$pmax2adj), col=2)
plot(subres$ACME, -log10(subres$pmax2adj), pch=".", main="genes_dn"); points(subres[subres$gene%in%genes_dn,]$ACME, -log10(subres[subres$gene%in%genes_dn,]$pmax2adj), col=3)
plot(subres$ACME, -log10(subres$pmax2adj), pch=".", main="peaks_up"); points(subres[subres$peak%in%peaks_up,]$ACME, -log10(subres[subres$peak%in%peaks_up,]$pmax2adj), col=4)
plot(subres$ACME, -log10(subres$pmax2adj), pch=".", main="peaks_dn"); points(subres[subres$peak%in%peaks_dn,]$ACME, -log10(subres[subres$peak%in%peaks_dn,]$pmax2adj), col=5)

plot(res$a, res$c, pch=".")
points(subres$a, subres$c, col=2)

idx_1 = subres$a >=0 & subres$c >=0
idx_2 = subres$a >=0 & subres$c < 0
idx_3 = subres$a < 0 & subres$c >=0
idx_4 = subres$a < 0 & subres$c < 0

bw = 10000
layout(matrix(1:6,2), respect=TRUE)
plot(density(res$distance            , bw=bw))
plot(density(subres$distance         , bw=bw))
plot(density(subres[idx_1, ]$distance, bw=bw))
plot(density(subres[idx_2, ]$distance, bw=bw))
plot(density(subres[idx_3, ]$distance, bw=bw))
plot(density(subres[idx_4, ]$distance, bw=bw))


idx_peaks_oi1 = unique(subres[idx_1, ]$peak)
idx_peaks_oi2 = unique(subres[idx_2, ]$peak)
idx_peaks_oi3 = unique(subres[idx_3, ]$peak)
idx_peaks_oi4 = unique(subres[idx_4, ]$peak)

length(idx_peaks_oi1);
length(idx_peaks_oi2);
length(idx_peaks_oi3);
length(idx_peaks_oi4);

tf_signal = apply(tf_scores[, idx_peaks_oi1], 1, median)
tf_bg = apply(tf_scores[, bg_peaks], 1, median)
enrich = sort(tf_signal - tf_bg, decreasing=TRUE)
head(enrich, 30)
tail(enrich, 30)

tf_signal = apply(tf_scores[, idx_peaks_oi2], 1, median)
tf_bg = apply(tf_scores[, bg_peaks], 1, median)
enrich = sort(tf_signal - tf_bg, decreasing=TRUE)
head(enrich, 30)
tail(enrich, 30)

tf_signal = apply(tf_scores[, idx_peaks_oi3], 1, median)
tf_bg = apply(tf_scores[, bg_peaks], 1, median)
enrich = sort(tf_signal - tf_bg, decreasing=TRUE)
head(enrich, 30)
tail(enrich, 30)

tf_signal = apply(tf_scores[, idx_peaks_oi4], 1, median)
tf_bg = apply(tf_scores[, bg_peaks], 1, median)
enrich = sort(tf_signal - tf_bg, decreasing=TRUE)
head(enrich, 30)
tail(enrich, 30)



plot(res[!duplicated(res$gene),]$cp, -log10(res[!duplicated(res$gene),]$pv3), pch=".")
points(subres[!duplicated(subres$gene),]$cp, -log10(subres[!duplicated(subres$gene),]$pv3), col=2)



plot(density(subres$distance))
plot(subres$c, subres$ACME)
plot(subres$cp, subres$ACME)
plot(subres$cp, subres$a)
plot(subres$b, subres$a)


tf_signal = apply(tf_scores[, idx_peaks_oi], 1, median)
tf_bg = apply(tf_scores[, bg_peaks], 1, median)
enrich = sort(tf_signal - tf_bg, decreasing=TRUE)
head(enrich, 30)
tail(enrich, 30)

# layout(1, respect=TRUE)
# plot(enrich, main="TF enrichment in peaks of interest")
# # abline(v=c(30, length(enrich)-30))
# i = c(1:30, (length(enrich) - 30):length(enrich))
# points(i, enrich[i], col=2, pch=16)
# text(i, enrich[i], names(enrich)[i], pos=4)


# library(ggplot2)
# library(ggrepel)

# df <- data.frame(
#   tf = names(enrich),
#   enrich = as.numeric(enrich),
#   rank = seq_along(enrich)
# )

# df$highlight <- df$rank %in% c(1:30, (nrow(df)-29):nrow(df))



# ggplot(df, aes(x = rank, y = enrich)) +
#   geom_line() +
#   geom_point(data = df[df$highlight, ], color = "red") +
#   geom_text_repel(
#     data = df[df$highlight, ],
#     aes(label = tf),
#     size = 5,
#     col="red",
#      segment.colour="black",
#     max.overlaps = Inf
#   ) +
#   labs(
#     title = "TF binding motif enrichement in peaks of interest",
#     x = "Rank",
#     y = "Enrichment"
#   ) +
#   theme_minimal()




# for (i in 1:30) {
#   # dev.new()
#   idx_tf = names(enrich)[i]
#   layout(matrix(1:6,2), respect=TRUE)
#   plot(enrich, main="TF enrichment in mediator peaks")
#   abline(v=c(30, length(enrich)-30))
#   points(i, enrich[i], col=2, pch=16)
#   text(i, enrich[i], idx_tf, pos=4)
  
#   den =density(tf_scores[idx_tf,bg_peaks])
#   plot(den, col="grey", main=paste0(idx_tf, " TF binding score distr   ibution"))
#   den2 = density(tf_scores[idx_tf, med_peaks], bw=den$bw)
#   # lines(den2$x, den2$y*length(med_peaks) / ncol(tf_scores), col=2)
#   lines(den2$x, den2$y, col=2)
#   legend("topleft", lty=1, col=c("red", "grey"), c("mediator peaks", "background") )
#   q = quantile(tf_scores[idx_tf,med_peaks], .90)
#   abline(v=q)
  
#   tf_peaks = colnames(tf_scores)[tf_scores[idx_tf,]>q]
   
#   idx_res = res$peak%in%tf_peaks & res$gene%in%med_genes
#   den = density(res[idx_res,]$a)
#   plot (den,    main=paste0("'a' distr. in ", idx_tf, " peaks"))
#   lines(density(          res$a, bw=den$bw), col="grey")
#   den = density(res[idx_res,]$b)
#   plot (den,    main=paste0("'b' distr. in ", idx_tf, " peaks"))
#   lines(density(          res$b, bw=den$bw), col="grey")
#   den = density(res[idx_res,]$ACME)
#   plot (den,    main=paste0("'ACME' distr. in ", idx_tf, " peaks"))
#   lines(density(          res$ACME, bw=den$bw), col="grey")
#   den = density(res[idx_res,]$c)
#   plot (den,    main=paste0("'c' distr. in ", idx_tf, " peaks"))
#   lines(density(          res$c, bw=den$bw), col="grey")
# }




# pmax2adj_tresh = 0.05
# layout(1, respect=TRUE)
# sel = res$pmax2adj < pmax2adj_tresh
# sel = res$pmax2adj < pmax2adj_tresh & res$pv < 0.01
# # sel = res$pv1 < 0.01
# # plot(-log10(res$pmax2adj), col=(rownames(res) %in% idx)+1)
# plot(res$ACME, -log10(res$pmax2adj), col=sel+1, pch=".")
# custom_hm(res$ACME, -log10(res$pmax2adj), LOGCOUNT=TRUE)

# med_peaks = unique(res$peak[sel])
# med_genes = unique(res$gene[sel])
# bg_peaks = unique(res$peak)
# bg_genes = unique(res$gene)

# length(med_peaks)
# length(med_genes)

# tf_signal = apply(tf_scores[, med_peaks], 1, median)
# tf_bg = apply(tf_scores[, bg_peaks], 1, median)

# enrich = sort(tf_signal - tf_bg, decreasing=TRUE)

# head(enrich, 30)
# tail(enrich, 30)
```


##  Enrichement
```{r, eval=FALSE}
# experimentalcustom from there
# stop("EFN")

plot(density(tf_scores))

custom_hm(res$ACME, -log10(res$pmax2adj), LOGCOUNT=TRUE)



srtres = res[order(res$ACME),]
layout(matrix(1:2,2))
plot(srtres$ACME)
tf = "FOSL2"
# tf_scores[tf,]
foo = cumsum(tf_scores[tf, srtres$peak] * srtres$ACME)
plot(foo, pch=".")
abline(a=0, b=rev(foo)[1] / nrow(srtres), col=2)

layout(matrix(1:6,2), respect=TRUE)
tf = "FOSL2"
plot(density(tf_scores[tf,]))
q = quantile(tf_scores[tf,], .95)
abline(v=q)

pks = colnames(tf_scores)[tf_scores[tf,]>q]

srtres$peak
plot(density(which(res[order(res$ACME),]$peak%in%pks)))
plot(density(which(res[order(res$RNA),]$peak%in%pks)))
plot(density(which(res[order(res$a),]$peak%in%pks)))
plot(density(which(res[order(res$b),]$peak%in%pks)))

plot(srtres$ACME)
abline(v=which(srtres$peak%in%pks), col=adjustcolor(2, alpha.f=.001))
wt = wilcox.test(which(srtres$peak%in%pks), which(!srtres$peak%in%pks))
wt$p.value

pvs = epimedtools::monitored_apply(t(t(rownames(m))), 1, function(tf) {
  q = quantile(m[tf,][m[tf,]>-10], .95)
  pks = colnames(m)[!is.na(m[tf,]) & m[tf,]>=q]
  wt = wilcox.test(which(foo$peak%in%pks), which(!foo$peak%in%pks))
  wt$p.value
})
names(pvs) = rownames(m)
pvs = sort(unlist(pvs))


plot(sort(-log10(unlist(pvs))))

layout(matrix(1:32,4), respect=TRUE)
for (i in 1:16) {
  print(i)
  tf = names(pvs)[i]
  q = quantile(m[tf,][m[tf,]>-10], .95)
  pks = colnames(m)[m[tf,]>=q]
  wt = wilcox.test(which(foo$peak%in%pks), which(!foo$peak%in%pks))  
  plot(foo$ACME, main=paste0(tf, " MW pv=", signif(wt$p.value, 3)), xlab="ranked peak:gene")
  abline(v=which(foo$peak%in%pks), col=adjustcolor(2, alpha.f=.01))
  den = density(which(foo$peak%in%pks)) 
  plot(den, col=2, xlim=c(0, nrow(foo)))
  lines(density(which(!foo$peak%in%pks)), col="grey", bw=den$bw)
  # if (tf %in% rownames(norm_counts_rna)) {
  #   md = lm(log2(norm_counts_rna[tf,]+1)~e[colnames(norm_counts_rna),]$age)
  #   plot(e[colnames(norm_counts_rna),]$age, log2(norm_counts_rna[tf,]+1), main=tf, xlab="age", ylab="log2(norm_counts_rna+1)")
  #   abline(md)
  # } else {
  #   plot.new()
  # }
}


# custom_hm(res$ACME, -log10(res$pmax2adj), LOGCOUNT=TRUE)
# peaks = peaks[rownames(peaks)%in%res$peak,]
# res = res[res$peak%in%rownames(peaks),]
# custom_hm(res$ACME, -log10(res$pmax2adj), LOGCOUNT=TRUE)
# custom_hm(res$ACME, peaks[res$peak, ]$len, LOGCOUNT=TRUE)

```

# Biblio

*The activity of early-life gene regulatory elements is hijacked in aging through pervasive AP-1-linked chromatin opening.*
Patrick R, Naval-Sanchez M, Deshpande N, Huang Y, Zhang J, Chen X, Yang Y, Tiwari K, Esmaeili M, Tran M, Mohamed AR, Wang B, Xia D, Ma J, Bayliss J, Wong K, Hun ML, Sun X, Cao B, Cottle DL, Catterall T, Barzilai-Tutsch H, Troskie RL, Chen Z, Wise AF, Saini S, Soe YM, Kumari S, Sweet MJ, Thomas HE, Smyth IM, Fletcher AL, Knoblich K, Watt MJ, Alhomrani M, Alsanie W, Quinn KM, Merson TD, Chidgey AP, Ricardo SD, Yu D, Jardé T, Cheetham SW, Marcelle C, Nilsson SK, Nguyen Q, White MD, Nefzger CM. 
**Cell Metab**. 2024 Aug 6;36(8):1858-1881.e23. doi: 10.1016/j.cmet.2024.06.006. Epub 2024 Jul 2. PMID: 38959897.

https://www.cell.com/cell-metabolism/fulltext/S1550-4131(24)00231-6



```{r plot_medres, eval=FALSE}

idx_ugenes = !duplicated(res$gene)
idx_upeaks = !duplicated(res$peak)

layout(matrix(1:3, 1), respect=TRUE)
plot(res[idx_ugenes,]$c, -log10(res[idx_ugenes,]$pv)  , col=adjustcolor(1, alpha.f=.2), xlab="Total Effect")
plot(res$cp, -log10(res$pv3), col=adjustcolor(1, alpha.f=.2), xlab="ADE", pch=".")
plot(res$c, res$cp, xlab="total Effect", ylab="ADE", pch=".", col=adjustcolor(1, alpha.f=.2))
abline(a=0, b=1, col=2)
abline(v=0, h=0, col="grey", lty=2)


layout(matrix(1:2, 1), respect=TRUE)
plot(res[idx_upeaks,]$a, -log10(res[idx_upeaks,]$pv1), xlab="a", ylab="-log10(pv_a)", col=adjustcolor(1, alpha.f=.2))
plot(res$b, -log10(res$pv2), xlab="b", ylab="-log10(pv_b)", col=adjustcolor(1, alpha.f=.2), pch=".")


layout(1, respect=TRUE)
idx = !is.na(res$pmax2) & res$pmax2adj < 0.05
# plot(-log10(res$pmax2adj), col=(rownames(res) %in% idx)+1)
plot(res$a*res$b, -log10(res$pmax2adj), col=idx+1, pch=".")
idx2 = res$gene == "EBF1"
points(res[idx2,]$a*res[idx2,]$b, -log10(res[idx2,]$pmax2adj), col=4, pch=1)
idx3 = idx & idx2
text(res[idx3,]$a*res[idx3,]$b, -log10(res[idx3,]$pmax2adj), res[idx3,]$peak)

# custom_hm(res$a*res$b, -log10(res$pmax2adj), col=idx+1, pch=".")

layout(matrix(1:2, 1), respect=TRUE)
plot(res$a * res$b , res$c, pch=".", xlab="ACME", ylab="Total Effect")
abline(a=0, b=1, col="grey", lty=2)
points(res[idx2,]$a * res[idx2,]$b , res[idx2,]$c, col=4)

plot(res$a * res$b , res$cp, pch=".", xlab="ACME", ylab="ADE")
abline(a=0, b=1, col="grey", lty=2)
abline(v=0, h=0, col="grey", lty=2)
points(res[idx2,]$a * res[idx2,]$b , res[idx2,]$cp, col=4)


# layout(1, respect=TRUE)


# pf$pvmax2adj = 0
# pf[idx, ]$pvmax2adj = 1
# hm_genome2(pf, "pvmax2adj", main="pvmax2adj")
```

```{r pairs, eval=FALSE}
genes = c("EBF1", "GABRB2", "NIPAL4", "HMMR")
pairs(t(log2(norm_counts_rna[genes,])))
genes = c("SGCD", "CCNJL")
pairs(t(log2(norm_counts_rna[genes,])))
```







# Session Information

```{r results="verbatim"}
JASPARPEAKS = COUNTPEAKS = MEDRES = FALSE
sessionInfo()
```



