---
title: "PCA on peaks"
author: "Florent Chuffart"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
---

```{r, echo=FALSE, eval=TRUE}
knitr::opts_chunk$set(collapse=TRUE, comment = "#>", fig.width=9, fig.height=6, eval=TRUE, echo=FALSE, results="hide")
```

# Pipeline Initialization

This section initializes the environment required for the analysis:

- `common.R` contains the utility functions used throughout the pipeline.
- `config` contains the dataset-specific parameters used by the script, such as the project name, dataset name, and species.
- The `COUNTPEAKS` variable controls whether the read-counting step using `featureCounts` is performed and is set to `FALSE` by default.
- The `peaks` prefix is used to reference the files generated by the pipeline.


```{r}
source("common.R")
source("config")
if (!exists("COUNTPEAKS")) {COUNTPEAKS = FALSE}
prefix = "mrg1.00e-5_peaks"
```





# Samples

This section defines the experimental design and associates each sample with its corresponding sequencing files:

- `samples.R` contains the sample-specific information and defines the `samples` variable used throughout the pipeline.
- The `e` data frame defines the **experimental grouping**, with one row per sample and columns pointing to the corresponding BigWig (`bw`) and BAM (`bam`) files and factors if needed.
- The `gsm` column stores the GEO sample like identifiers, which are also used as row names to uniquely identify each sample.
- The file paths are constructed dynamically from the dataset (`gse`), sample name (`samples`), species, annotation, genome version, and sequencing configuration.

```{r label="samples"}
source("samples.R")
e = data.frame(
  gsm = samples,
  bam = paste0("/home/chuffarf/projects/datashare/", gse, "/", samples, "_end-to-end_trimno_bowtie2_", species, "_", annotation, "_", version, "_srt.bam"),
  bw = paste0("/home/chuffarf/projects/datashare/", gse, "/", samples, "_end-to-end_trim30_bowtie2_", species, "_", annotation, "_", version, "_srt_", sr_or_pe, "_30_4_RPKM.bw")
)
rownames(e) = e$gsm

if (file.exists("exp_grp.R")) {
  source("exp_grp.R")
}

head(e)
dim(e)
```





# Peaks


```{r peaks, fig.height=6}
# #	Nom de la colonne	Description	Type / Exemple
# 1	chrom	Nom du chromosome	chr1
# 2	chromStart	Position de début du pic (0-based, inclusif)	3455000
# 3	chromEnd	Position de fin du pic (1-based, exclusif)	3455500
# 4	name	Nom du pic	peak_00001
# 5	score	Score de 0 à 1000 (utile pour visualisation UCSC)	500
# 6	strand	Brin (+, -, ou . si non applicable)	.
# 7	signalValue	Enrichissement du signal au sommet du pic	12.45
# 8	pValue	–log10(p-value) de la significativité du pic	7.30
# 9	qValue	–log10(q-value, FDR ajustée)	5.21
# 10	peak	Position du sommet du pic par rapport à chromStart (en bp). -1 si inconnu.	250
peaks = read.table(paste0("macs3_output_merged/", prefix, ".narrowPeak"))
rownames(peaks) = peaks[,4]
colnames(peaks) = c("chrom", "chromStart", "chromEnd", "name", "score", "strand", "signalValue", "pValue", "qValue", "peak")
peaks$len = peaks$chromEnd - peaks$chromStart
peaks = peaks[nchar(peaks$chrom)<=5,]
head(peaks)
dim(peaks)

# Filtering peaks
head(peaks)
dim(peaks)
peaks$key = paste0(peaks[[1]], ":", peaks[[2]], "-", peaks[[3]])
length(unique(peaks$key))
# qv_thresh = peaks[order(peaks$qValue, decreasing=TRUE),]$qValue[40000]
# peaks = peaks[peaks$qValue>=qv_thresh,]
dim(peaks)
length(unique(peaks$key))
peaks = peaks[!duplicated(peaks$key),]
dim(peaks)
length(unique(peaks$key))

# Stat peaks
custom_hm(peaks$qValue , peaks$pValue     , xlab="qValue", ylab="pValue     ", LOGCOUNT=TRUE, main=prefix)#, xlim=c(0, quantile(peaks$qValue, .99), ylim=c(0, quantile(peaks$pValue     , .99))))
custom_hm(peaks$qValue , peaks$signalValue, xlab="qValue", ylab="signalValue", LOGCOUNT=TRUE, main=prefix)#, xlim=c(0, quantile(peaks$qValue, .99), ylim=c(0, quantile(peaks$signalValue, .99))))
custom_hm(peaks$len    , peaks$signalValue, xlab="len   ", ylab="signalValue", LOGCOUNT=TRUE, main=prefix)#, xlim=c(0, quantile(peaks$len   , .99), ylim=c(0, quantile(peaks$signalValue, .99))))
custom_hm(peaks$len    , peaks$qValue     , xlab="len   ", ylab="qValue     ", LOGCOUNT=TRUE, main=prefix)#, xlim=c(0, quantile(peaks$len   , .99), ylim=c(0, quantile(peaks$qValue     , .99))))
# stop("EFN")
```

```{r prefix2}
prefix2 = paste0(prefix, "_", prefix)
prefix2 = paste0(prefix)
```



## Counts




## Peak counts







```{r count_peaks, eval=COUNTPEAKS}
# stop("EFN")
# Counts
feat_saf = peaks[, c(4,1:3, 6)]
# feat_saf = feat[1:100,c(4,1:3, 6)]
colnames(feat_saf) = c("GeneID", "Chr", "Start", "End", "Strand")
feat_saf$Start = as.numeric(feat_saf$Start)
feat_saf$End   = as.numeric(feat_saf$End  )
saf_filename = paste0(prefix2, ".saf")
write.table(feat_saf, file=saf_filename, sep="\t", quote=FALSE,row.names=FALSE, col.names=FALSE)


cmd = "featureCounts"
mmq = 30
# if PE
if (sr_or_pe=="PE") {
  opt_pe="-p"
} else {
  opt_pe=" "  
}
args = paste0(
  " ", opt_pe, " -a ", saf_filename, " -F SAF -s 0 -Q ", mmq, " -T 32 -o ", 2, "_count.txt ", 
  paste0(e$bam , " ", collapse=" "),
  ""
)
print(paste(cmd, args))
system2(cmd, args)
system2("cat", paste0(prefix2, "_count_.txt.summary"))
# stop("EFN")
```




##  Raw counts

```{r raw_counts_peaks}
raw_counts_summary = read.table(paste0(prefix2, "_count.txt.summary"), header=TRUE, row.names=1)
colnames(raw_counts_summary) = sub(paste0(".*", gse, "\\.([^.]+)_end\\.to\\.end.*"), "\\1", colnames(raw_counts_summary))
layout(matrix(1:2, 1), respect=TRUE)
# barplot(table(as.matrix(raw_counts_summary), margin=2), las=2)
barplot(prop.table(as.matrix(raw_counts_summary), margin=2), las=2)

raw_counts = read.table(paste0(prefix2, "_count.txt"), header=TRUE)
rownames(raw_counts) = raw_counts[,1]
pf = raw_counts[,1:6]
options(scipen=999)
write.table(peaks[rownames(pf) ,1:6]  , file=paste0(prefix2, "peaks.bed"), sep="\t", quote=FALSE, row.names=FALSE, col.names=FALSE)
options(scipen=0)

raw_counts = as.matrix(raw_counts[,-(1:6)])
colnames(raw_counts) = sub(paste0(".*", gse, "\\.([^.]+)_end\\.to\\.end.*"), "\\1", colnames(raw_counts))
head(raw_counts[,1:3])
tail(raw_counts[,1:3])
dim( raw_counts)
```

## CP50M norm counts

```{r}
```

## Deseq2 norm counts


```{r norm_counts_peaks, eval=TRUE}
if (!all(e$gsm %in% colnames(raw_counts))) {stop("problem gsm vs. colnames(raw_counts)")}
countData = raw_counts[,e$gsm]
colnames(countData) = rownames(e)
colData = e[colnames(countData), ]
# Vérifie que les noms de colonnes correspondent
if (!all(colnames(countData) == rownames(colData))) {stop("probleme data vs. e")}

dds <- DESeq2::DESeqDataSetFromMatrix(
  countData = countData,
  colData = colData,
  design = ~ 1
)

# dds <- DESeq2::DESeq(dds)
# res_age <- DESeq2::results(dds, name = "age")

dds = DESeq2::estimateSizeFactors(dds)

norm_counts_peaks = DESeq2::counts(dds, normalized=TRUE)
head(norm_counts_peaks[,1:3])
dim(norm_counts_peaks)
```


# PCA


```{r pca_norm_counts_peaks}
data = log2(as.matrix(t(norm_counts_peaks))+1)
pca = prcomp(data, scale=FALSE)
v = pca$sdev * pca$sdev
p = v / sum(v) * 100
# # --- MCLUST sur les scores PCA ---
# library(mclust)
# mc = mclust::Mclust(pca$x[,2:6], G=5)
# cl = mc$classification
# col = cl

layout(matrix(1:6,2), respect=TRUE)
# layout(matrix(1:2,1), respect=TRUE)
barplot(p, ylab="% of variance explained", xlab="components")

if (exists("factoi")) {
  if (is.factor(e[,factoi[1]])) {
    col = as.numeric(e[rownames(pca$x),factoi[1]])
    legend("topright", col=1:length(levels(e[,factoi[1]])), levels(e[,factoi[1]]),pch=16)   
  } else if (is.logical(e[,factoi[1]])) {
    col = as.numeric(e[rownames(pca$x), factoi[1]]) + 1
    legend("topright", col=1:2, levels(e[,factoi[1]]), c("No", "Yes"), pch=16)
  }
} else {
  col = rep(1, nrow(pca$x))
}
names(col) = rownames(pca$x)

for (i in 1:5) {
  j = i+1
  plot(
    pca$x[,i], pca$x[,j],
    xlab=paste0("PC", i, "(", signif(p[i], 3), "%)"),
    ylab=paste0("PC", j, "(", signif(p[j], 3), "%)"),
    col=col,
    pch=16
  )
}

layout(1, respect=TRUE)

i=1
j = i+1
plot(
  pca$x[,i], pca$x[,j],
  xlab=paste0("PC", i, "(", signif(p[i], 3), "%)"),
  ylab=paste0("PC", j, "(", signif(p[j], 3), "%)"),
  col=adjustcolor(col, alpha.f=0.5),
  pch=16
)

text(pca$x[,i], pca$x[,j], rownames(pca$x))

```


# Fingerprint

```{r}
data = as.matrix(t(raw_counts))
# col = as.numeric(e[,factoi[1]])
# names(col) = rownames(e)

layout(1, respect=TRUE)
plot(0,0, col=0, xlim=c(1, ncol(data)), ylim=0:1, xlab="Ranked genomic peaks  ", ylab="Cumulative fraction of signal")
sapply(rownames(data), function(k){
  l = data[k,]
  tot = sum(sort(l))
  cs = cumsum(sort(l))
  lines(cs/tot, type="l", col=adjustcolor(col[k], alpha.f=0.7))
}) 
```



# Session Information

```{r results="verbatim"}
COUNTPEAKS = FALSE
sessionInfo()
```



























